---
title: "salmon outmigrant phenology"
author: "Pete Nelson"
format: 
  html:
    embed-resources: true
execute:
  execute-dir: project
editor: visual
---

# introduction

The intent here is to explore the beach seine data from the DJFMP dataset for patterns in the presence and abundance of juvenile Chinook salmon in the lower Sacramento River and the Delta. These data were first collected in 1976 and the most recent data are from water year 2024.

# setup

One package *not* available on CRAN: If necessary, get it using `remotes::install_github("doi-usgs/EflowStats@v5.2.0")`

```{r}
#| label: packages
#| message: false

# library(conflicted) # causes WAAAY more trouble than it solves!

library(here)
library(readxl)
library(janitor) 
library(ggplot2)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
library(EflowStats)
library(fishualize)
library(tidybayes)
library(nlme)
library(tidyverse)

```

# load data

```{r}
#| label: seine-data
#| echo: true

# beach seine data ('raw' includes all spp)
raw <- 
  read_rds(here("salmon/data/raw/raw.rds"))

# naming conventions
taxonomy <- 
  read_rds(
  here("salmon/data/raw/dt4.rds"))

# site information
sample_sites <- 
  read_rds(
    here("salmon/data/raw/dt5.rds"))

# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

```

## storm data

Let's pull in Rosie's storm data including *first flush* events, both timing and magnitude.

```{r}
#| label: storm-data
#| message: false
#| warning: false


load(here("data/processed/storms/StormData.RData"))

# use 'Firststorms'
storms <-
  Firststorms %>% 
  arrange(WY) %>% 
  # wyd of first flush events for each wy
  mutate(wyd = get_waterYearDay(Date))

write_rds(storms, here("salmon/data/processed/storms.rds"))

```

# wrangle data

## choose sites

There are beach seine data from a large number of sites, but the consistency with which these sites were samped varies widely. In a separate document, I provide the details used to select sites that were sampled consistently from water year 1976-2024. Even at these sites, however, there are missing data, particularly for the months April-December prior to 1992.

```{r}
#| label: limit-sites
#| message: false
#| warning: false


# list of beach seine stations identified minimally adequate for 
# temporal coverage (tibble includes region and region_label)
sites <- 
  read_rds(here("salmon/data/processed/sites.rds"))

# select data from those minimum effort sites and for complete water years (1976-10-01 onwards)
dat <-
  raw %>% 
  filter(station %in% sites$station,
         # limit to data collected during complete water years
         date >= "1976-10-01") %>% 
  mutate(year = year(date),
         month = month(date)) %>% 
  relocate(region, .after = id_code) %>% 
  relocate(year:month, .before = date)

# create monthly coverage indicator and retain region_code
coverage <- dat %>%
  distinct(station, year, month, region) %>%
  # 'tag' months sampled
  mutate(sampled = 1)

```

## water year types

In California, *water year types* are classifications (Wet, Above Normal, Below Normal, Dry, and Critical) that are used to categorize hydrological conditions over a 12-month period (October 1–September 30), based, for the Sacramento and San Joaquin valleys, on unimpaired runoff. These designations are used for making water management decisions and probably are not the best environmental measures available for understanding salmon ecology, but they are used by managers and offer a reasonable starting point for exploring the relationship between environment and salmon migration. Additional environmental metrics will be explored in the future.

Add data on *water year types* to the data:

```{r}
#| label: wy-wyd
#| message: false
#| warning: false


# add variables: water year and day of water year
# select data begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

dat <-
  dat %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = dataRetrieval::calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = EflowStats::get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  dplyr::select(c(station:water_temp, organism_code, fork_length, race_by_length, count)) %>% 
  # add 'sample_code' to identify each unique seine event
  group_by(date, time, station) %>% 
  mutate(sample_id = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(sample_id) %>% 
  write_rds(., here("salmon/data/raw/dat.rds"))

```

# analysis

## accumulation curves

Calculate the accumulated number of salmon over the course of each water year and aggregate over selected beach seine sites. Determine a cumulative percentage for each day over the entire year and for the 3-month period January, February, and March. Plot annual accumulation curves to show the pattern of juvenile outmigration for each water year, as well as for each January-March period for each year. The truncated data are used as these months were sampled consistently throughout the time series; sampling effort during the remaining months, April-December was limited prior to 1992.

```{r}
#| label: counts
#| message: false
#| warning: false


# daily counts
# pivot to simplify summing Chinook counts (including 0's) for each sample unit
seine <-
  dat %>% 
  pivot_wider(
    # relies on 'sample_id' to distinguish among sample units, but retains environmental data
    id_cols = c(sample_id:time, water_temp),
    # columns for all fish spp (only CHN needed for this analysis)
    names_from = organism_code,
    # log number of each speciesxfork_length for each seine (sample_id)
    values_from = count,
    # if no matches, enter '0'
    values_fill = 0,
    # sum counts if multiple matches
    values_fn = sum
  ) %>% 
  dplyr::select(c(sample_id:water_temp, CHN)) %>% 
  left_join(sites, by = "station") %>% 
  mutate(location = location.x, .keep = "unused") %>% 
  dplyr::select(-c(region, location.y, method_code)) %>% 
  relocate(c(region_label, location, river_km), .after = station)

### daily counts -----
# aggregate to daily totals bc freq >>1 seine per day

# aggregated across all selected sites
daily_delta <-
  seine %>%
  group_by(wy, month, wd) %>%
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  )

# aggregated by region
daily_regional <-
  seine %>% 
  group_by(region_label, wy, month, wd) %>% 
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent------
# limited to the period from Jan 1-Mar 31
# calc cumulative percentage for each wy during that 3-month period, delta-wide
delta_cumul <-
  daily_delta %>% 
  filter(month <= 3) %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/delta_cumul.rds"))

# cumulative totals and percentages for Jan-Mar by region
regional_cumul <-
  daily_regional %>% 
  filter(month <= 3) %>% 
  group_by(wy, region_label) %>% 
  arrange(region_label, wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, month, wd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/regional_cumul.rds"))

# cumulative totals and percentages for Jan-Mar by station
station_cumul <-
  seine %>% 
  group_by(station, wy, month, wd, river_km) %>% 
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  filter(month <= 3) %>% 
  group_by(wy, station) %>% 
  arrange(station, wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(station, wy, month, wd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/station_cumul.rds"))

```

Plot the accumulation curves.

```{r}
#| label: annual-accumulation
#| message: false
#| warning: false

# annual (whole year) accumulation curves, lumping data from all sites
daily_delta %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ggplot(aes(x = wd,
           y = cumul_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  xlim(0, 300) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  # highlight 1977 data
  annotate("text", x = 250, y = 30,
           label = "1977", color = "red", fontface = "bold") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Delta-Wide Accumulation Curves",
       subtitle = "data from 1977-2024",
       color = "water year") +
  theme_bw()

```

```{r}
#| label: 3month-accumulation
#| message: false
#| warning: false


regional_cumul %>% 
  arrange(region_label, wy, wd) %>%
  mutate(region_label = fct_relevel(region_label, 
                                     "Lower Sacramento River", "North Delta", 
                                     "Central Delta", "South Delta")) %>% 
  ggplot(aes(x = wd,
             y = cumul_percent,
             color = wy,
             group = wy)) +
  geom_line() +
  xlim(50, 200) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of 3-Month Total", 
       title = "Juvenile Chinook Regional Accumulation Curves",
       subtitle = "January - March, 1977-2024",
       color = "Water Year") +
  facet_wrap(~region_label, nrow = 2) +
  theme_bw()

```

```{r}
#| label: sta-accumul

station_cumul %>% 
  filter(!is.nan(cumul_percent)) %>% 
  arrange(station, wy, wd, river_km) %>%
  mutate(station = as.factor(station)) %>% 
  ggplot(aes(x = wd,
             y = cumul_percent,
             color = river_km,
             # group = wy
             )) +
  geom_line() +
  xlim(50, 200) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of 3-Month Total", 
       title = "Juvenile Chinook Accumulation Curves by Station",
       subtitle = "January - March, 1977-2024",
       color = "River KM") +
  theme_bw()

```

## benchmarks

Determine the day of each water year when 10%, 50% and 90% of the total annual numbers are reached.

```{r}
#| label: benchmarks
#| message: false

## ID benchmarks ----
# day of water year when the 10%, 50% and 90% of total annual numbers--accumulation benchmarks

# for entire year, entire delta
annual_benchmarks <-
  daily_delta %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

three_month_delta_benchmarks <-
  delta_cumul %>% 
  arrange(wy, wd) %>% 
  group_by(wy) %>% 
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

three_month_regional_benchmarks <-
  regional_cumul %>% 
  arrange(region_label, wy, wd) %>% 
  group_by(region_label, wy) %>% 
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

three_month_station_benchmarks <-
  station_cumul %>% 
  arrange(station, wy, wd) %>% 
  group_by(station, wy) %>% 
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

```

## water year type

Add the water year type data (loaded previously) to the percentile data for analysis and graphics.

```{r}
#| label: wy-type
#| message: false
#| warning: false


## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

three_month_regional_benchmarks <-
  wy_type %>% 
  # select Sac watershed; exclude SJR watershed
  filter(watershed == "Sacramento Valley") %>% 
  # merge with benchmark data
  right_join(three_month_regional_benchmarks,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, 
                           c("C", "D", "BN", "AN", "W")),
         # center wy to facilitate model convergence
         wy_meanCentered = wy - mean(wy)) %>% 
  relocate(wy_meanCentered, .after = wy)

# make 'wyt' an ordered factor
three_month_regional_benchmarks <- 
  three_month_regional_benchmarks %>%
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      # must be low to high, where C<D<BN<AN<W
                      ordered = TRUE)) %>% 
  write_rds(
    here("salmon/data/processed/three_month_regional_benchmarks.rds"))


```

## migration timing

Plot the water year days when each benchmark (10, 50, or 90%) is reach for each water year.

```{r}
#| label: wy-effect
#| message: false
#| warning: false


# quick data qaqc check (shouldn't be any NAs)
unique(three_month_regional_benchmarks$region_label)

# 10% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_10 benchmark
  filter(percentile == "day_10") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing, 10% Benchmark",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "10% of 3-Month Total Reached (day of water year)")

ggsave(here("salmon/plots/benchmark_10pct.png"))

# 50% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_50 benchmark
  filter(percentile == "day_50") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing, 50% Benchmark",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "50% of 3-Month Total Reached (day of water year)")

ggsave(here("salmon/plots/benchmark_50pct.png"))

# 90% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_90 benchmark
  filter(percentile == "day_90") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing, 90% Benchmark",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "90% of 3-Month Total Reached (day of water year)")

ggsave(here("salmon/plots/benchmark_90pct.png"))


```

## models

### change over time and water year type effects

Testing for potential effects of water year (i.e., change over time), water year type, regional differences, and differences in the three benchmarks (is there a difference in the slopes).

```{r}
#| label: bayes1
#| message: false
#| warning: false

# ensure that we've got a nested directory
# output_dir <- here("salmon", "results")
# if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# I've used a Gaussian distribution (at least for now) because the distributions 
# for each benchmark look "normalish." See
three_month_regional_benchmarks %>%
  ggplot(aes(wd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")
# I'll do a pp_check of the model; if this shows the model giving wd=60 or wd=250 with
# non-trivial probabilities, I'll have to swtich to Beta or truncated normal distribution

bmod1 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = gaussian,
      wd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      # set seed
      seed = 42,
      # brms automatically appends '.rds'
      file = here("salmon/results/bmod1"),
      file_refit = "on_change"
      )

pp_check(bmod1) # looks good!
(bmod1_sum <- summary(bmod1)) # see analysis
plot(bmod1) # looks good
conditional_effects(bmod1, "wy_meanCentered:percentile") # plot doesn't add much

```

All `Rhat = 1.00`, so the model converged cleanly. We have effective sample sizes in the low thousands and `n = 507`, so we can take the estimates seriously.

Without going into the output details here (I will with the best model), the entire distribution of the `wd` benchmarks (10th, 50th, 90th percentiles) is moving more than 0.4 days earlier each year, and the shift is statistically credible. The three benchmarks are shifting in parallel rather than diverging or converging. Fish arrive in the Lower Sacramento River about a week prior to their arrival in the Delta, and timing to reach sites within the Delta are indistinguishable from each other. Arrival times are slightly earlier during wet years than in dry years, but that gap has narrowed over time. However, water year hydrology is genuinely heavy-tailed — ENSO events and atmospheric rivers can produce years that are extreme outliers relative to the typical range (see the accumulation curves for 1977 above for example), and I don't want those years unduly affecting my coefficient estimates. A Gaussian model will try harder to accommodate outliers by distorting the fitted mean, whereas a Students t-distribution is more robust to these extreme values and less likely to overfit the model.

I ran the same model using a Students t-distribution (`bmod2`, below):

```{r}
#| label: robust
#| message: false
#| warning: false

bmod2 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = student,
      wd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod2"),
      file_refit = "on_change"
      )

pp_check(bmod2) # looks good!
(bmod2_sum <- summary(bmod2)) # see analysis
plot(bmod2) # looks good
conditional_effects(bmod2, "wy_meanCentered:percentile") # doesn't add much

```

This a real improvement over `bmod1` with the `nu` estimate \<10. I compare these two models directly later in this report, but first let's start looking at the possible effects of first flush events.

### first flush effect

Pull in Rosie's storm data and integrate that with the benchmark data from above.

```{r}
#| label: ff-data
#| message: false
#| warning: false


storms <- 
  read_rds(here("salmon/data/processed/storms.rds")) %>% 
  mutate(ff_wyd = wyd,
         .keep = "unused")

# no first flush storms for 1977, 2020, 2022 and 2025--filter(is.na(WY))

ff_data <- 
  three_month_regional_benchmarks %>% 
  left_join(storms,
            join_by(wy == WY),
            keep = FALSE) %>% 
  dplyr::select(wy, wy_meanCentered, wyt, ff_wyd, percentile, wd, SAC, region = region_label) %>% 
  filter(!is.na(ff_wyd))

```

Let's see if there are any patterns by plotting some of these data.

```{r}
#| label: plots
#| message: false
#| warning: false


ff_data %>%
  ggplot(aes(wd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")
# looks good; skew doesn't appear excessive

ff_data %>% 
  ggplot(aes(x = ff_wyd, y = wd, color = percentile)) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(title = "First Flush and Chinook Outmigration Timing",
       subtitle = "outmigration during Jan - Mar only",
       x = "First Flush Timing (day of water year)",
       y = "% Accumulation Benchmark (day of water year)") +
  theme_bw()

ggsave(here("salmon/plots/first_flush_timing.png"))

```

### first flush timing

The migration data look cool, but first, does the timing of first flush events change over time? Plotting the timing and then a simple linear regression suggests that first flush events have come later over this time period.

```{r}
#| label: ff-timing


# does the timing of the ff events change over time?
ff_data %>% 
  filter(percentile == "day_10", region == "Central Delta") %>% 
  ggplot(aes(x = wy, y = ff_wyd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Day of First Flush",
       subtitle = "1978 - 2024",
       x = "Water Year",
       y = "Day of First Flush (day of water year)") +
  theme_bw()

lm_ff_annual <-
  lm(data = ff_data %>% 
       # ensure we aren't pseudo-replicating
       filter(percentile == "day_10", region == "Central Delta"), 
     ff_wyd ~ wy)

summary(lm_ff_annual) # no. there's a slope but way noisy

# evidence of autocorrelation?
acf(residuals(lm_ff_annual))
# nope

# what's the potential role of water year type here, given that it's almost
# certainly correlated with the timing of first flush events?
lm_ff_wyt <- lm(
  data = ff_data %>% 
    filter(percentile == "day_10", region == "Central Delta"),
  ff_wyd ~ wy + wyt
)
summary(lm_ff_wyt)

# let's compare these models
AIC(lm_ff_annual, lm_ff_wyt)
anova(lm_ff_annual, lm_ff_wyt)

```

The plot suggested that *maybe* these events were happening later over time, but the regression analysis shows that, while there is an effect of water year *type* on the timing of first flush, there's insufficient evidence to suggest that there's been a change over time. The improvement to the model with the addition of `wyt` is modest (p=0.056), but seems real to me.

### first flush effects on migration timing

Now the question is, does the timing of first flush events affect the timing of juvenile salmon outmigration, as the figure relating migration timing to the timing of first flush events above suggests? I expect that *when* we get that first flush will affect the timing for that 10% benchmark, but that the effect will be reduced or absent with regard to reaching the 50 and 90% benchmarks. Note that we don't have juvenile salmon accumulation data from South Delta for years 1989, 1991, 1993, 1997, 2009, 2010, 2012-2016, 2018, 2019, 2022, and 2023.

```{r}
#| label: ff-models
#| message: false
#| warning: false

bmod3 <- 
  brm(data = ff_data,
      # 1st interaction lets slope of wd differ across benchmarks
      # I expect that day_10 will track ff closely but day_90 not so much
      wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered,
      # student-t helped in bmod2, so
      family = student,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod3"),
      file_refit = "on_change")

pp_check(bmod3) # looks good!
(bmod3_sum <- summary(bmod3)) # nu=7.21; student distribution is helping
plot(bmod3) # looks good
conditional_effects(bmod3, "ff_wyd:percentile") # ff matters more early in migr

bmod4 <- 
  brm(data = ff_data,
      # within yr obs may not be independent; this may help
      wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy),
      family = student,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod4"),
      file_refit = "on_change")

pp_check(bmod4) # looks good!
(bmod4_sum <- summary(bmod4)) # regional effects emerge more clearly
plot(bmod4) # looks good
conditional_effects(bmod4, "ff_wyd:percentile") # dividing the migration into
# early, mid and late allows us to see there's a real effect of the ff timing
# albeit primarily on those first, early migrants

# define some priors based earlier model output
my_priors <- c(
  prior(normal(150, 50), class = Intercept),
  prior(normal(0, 1), class = b, coef = ff_wyd),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_50"),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_90"),
  prior(normal(0, 2), class = b, coef = wy_meanCentered),
  prior(normal(0, 30), class = b), # now only hits percentile & region contrasts
  prior(normal(0, 20), class = sd),
  prior(exponential(1), class = sigma),
  prior(gamma(2, 0.1), class = nu)
)

bmod5 <- brm(
  data = ff_data,
  wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy),
  family = student,
  prior = my_priors,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 42,
  file = here("salmon/results/bmod5"),
  file_refit = "on_change"
)

pp_check(bmod5) # looks GREAT
pp_check(bmod5, type = "stat", stat = "mean")
(bmod5_sum <- summary(bmod5)) # what's going on w S Delta? SJR fish?!
plot(bmod5) # looks good
conditional_effects(bmod5, "ff_wyd:percentile")

```

The diagnostics look clean again — all `Rhat` = 1.00, ESS well above 1000, and `nu` well below 10. The results look really good, but I think we should add water year type to the model to see if overall water year conditions affect migration timing. This could, for example, affect the tail end of the migration (`day_90`) while having little or no effect on the first 10%.

```{r}
#| label: ff+wyt

# we're going to have to revise the priors though:
get_prior(
  wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + mo(wyt) + (1 | wy),
  data = ff_data,
  family = student
) # this is mostly to ensure that I'm using the right terms

my_priors_bmod6 <- c(
  prior(normal(150, 50), class = Intercept),
  prior(normal(0, 1), class = b, coef = ff_wyd),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_50"),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_90"),
  prior(normal(0, 2), class = b, coef = wy_meanCentered),
  prior(normal(0, 30), class = b, coef = mowyt),
  prior(normal(0, 30), class = b), # percentile & region contrasts
  prior(normal(0, 20), class = sd),
  prior(exponential(1), class = sigma),
  prior(gamma(2, 0.1), class = nu),
  prior(dirichlet(1), class = simo, coef = mowyt1)
)

bmod6 <- brm(
  data = ff_data,
  # note addition of the monotonic wyt
  wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + mo(wyt) + (1 | wy),
  # sample_prior = "only", # run pp_check(bmod6); if ok run without
  family = student,
  prior = my_priors_bmod6,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 42,
  file = here("salmon/results/bmod6"),
  file_refit = "on_change"
)

pp_check(bmod6)
pp_check(bmod6, type = "stat", stat = "mean")
(bmod6_sum <- summary(bmod6)) # what's going on w S Delta? SJR fish?!
plot(bmod6) # looks good
conditional_effects(bmod6, "ff_wyd:percentile")

bmod5 <- add_criterion(bmod5, "loo")
bmod6 <- add_criterion(bmod6, "loo")
loo_compare(bmod5, bmod6)

```

The model including water year type (`bmod6`) showed negligible improvement in predictive accuracy over the simpler model, `bmod5`, (Δelpd = –0.3 ± 0.2), and the water year type coefficient was not credibly different from zero, so we should go with the more parsimonious `bmod5`.

Here's a detailed review of the bmod5 output:

- `Intercept` (101.26, CI \[93.36-109.29\]): This is the expected `wd` for the 10th percentile benchmark in the reference region (Central Delta) at the mean year when `ff_wyd` = 0. Since day 0 of the water year isn't meaningful biologically, treat this more as a mathematical anchor than an interpretable quantity.

- Day of the first flush (`ff_wyd`; 0.21, CI \[0.13-0.29\]): Credibly positive — each day later that first flush occurs, the 10th percentile outmigration benchmark shifts \~0.21 days later. First flush timing matters: a pulse that comes a month late pushes early migrants roughly 6 days later.

- Interactions between first flush timing and the benchmarks:

  - 10th percentile: **+0.21** (the main effect)

  - 50th percentile: 0.21 – 0.15 = **+0.06** (barely a relationship)

  - 90th percentile: 0.21 – 0.23 = **-0.02** (negligible)

- The timing of first flush events clearly affects when the first juvenile migrants reach the Delta. Presumably other factors (water temperature, cumulative flow conditions, individual variation) determine when the bulk of the outmigrants get there.

- Regional effects (`region`) are comparable to results from the first models. The lower Sacramento is about 8 days earlier than the reference region; fish reach the North Delta a little more than 2 days before the Central Delta. Timing for the South Delta can't be distinguished from the reference region. The latter result could be due to missing data for a large number of years, the effect of San Joaquin river fish entering the Delta independently, or both.

- Change over time (`wy_meanCentered`; -0.30, CI \[-0.47--0.12\]): There is a credible temporal trend towards earlier migration.

- sigma (8.56): Smaller than `bmod1` (14.62); confirms that first flush timing adds real explanatory power.

- Bottom line: There has been a shift in the timing of salmon outmigration over this 45 year period, with the first 10% of each year's cohort now reaching the Delta nearly two weeks earlier than the late 1970's. First flush timing is also a credible predictor of *early* outmigration (10th percentile) but not of bulk or late outmigration, as we hypothesized. Outmigration phenology appears to be changing over the long term (decades), but is also strongly affected by year-to-year differences.

Compare model performance.

```{r}
#| label: pareto-k

models <- 
  list(bmod1 = bmod1, bmod2 = bmod2, bmod3 = bmod3, bmod4 = bmod4, bmod5 = bmod5, bmod6 = bmod6)

# fit loo for each
loos <- map(models, loo)

# summary table of k diagnostics per model
k_summary <- imap_dfr(loos, ~{
  k <- .x$diagnostics$pareto_k
  tibble(
    model     = .y,
    n_obs     = length(k),
    k_max     = max(k),
    n_ok      = sum(k <= 0.5),
    n_okish   = sum(k > 0.5 & k <= 0.7),
    n_bad     = sum(k > 0.7 & k <= 1.0),
    n_veryBad = sum(k > 1.0)
  )
})
print(k_summary)

# detailed table of every observation with k > 0.7
bad_points <- imap_dfr(loos, ~{
  k <- .x$diagnostics$pareto_k
  idx <- which(k > 0.7)
  tibble(model = .y, obs = idx, pareto_k = k[idx])
}) |>
  arrange(desc(pareto_k))

if (nrow(bad_points) == 0) {
  message("All Pareto k values <= 0.7 across all models.")
} else {
  message("Observations with Pareto k > 0.7:")
  print(bad_points)
}

```

```{r}
#| label: loo


# loo_compare(loos$bmod1, loos$bmod2, loos$bmod3, loos$bmod4)
# can't compare all four directly bc unequal sample sizes

loo_compare(loos[c("bmod1", "bmod2")])
loo_compare(loos[c("bmod3", "bmod4", "bmod5", "bmod6")])

```

Models `bmod2` and `bmod5` outperform their respective alternatives.

### interpretation

While output from the two models designed to study the importance of first flush timing on migration phenology are similar, `mbod5` outperformed `bmod3` and `bmod4` so let's work through the output for `mbod4`. The diagnostics look clean again — all Rhat = 1.00, ESS well above 1000. How should we understand these results?

- `Intercept` (101.26, CI \[93.36-109.29\]): This is the expected `wd` for the 10th percentile benchmark in the reference region (Central Delta) at the mean year when `ff_wyd` = 0. Since day 0 of the water year isn't meaningful biologically, treat this more as a mathematical anchor than an interpretable quantity.

- Day of the first flush (`ff_wyd`; 0.21, CI \[0.13-0.29\]): Credibly positive — each day later that first flush occurs, the 10th percentile outmigration benchmark shifts \~0.21 days later. First flush timing matters: a pulse that comes a month late pushes early migrants about 6 days later.

- Interactions between first flush timing and the benchmarks:

  - 10th percentile: **+0.21** (the main effect)

  - 50th percentile: 0.21 – 0.15 = **+0.06** (barely a relationship)

  - 90th percentile: 0.21 – -0.23 = **-0.02** (even less of a relationship)

- The timing of first flush (`ff_wyd`) events clearly impacts when the first juvenile migrants (`percentileday_10`) reach the Delta. Presumably other factors (water temperature, cumulative flow conditions, individual variation) determine when the bulk of the outmigrants get there (`percentileday_50` and `percentileday_90`).

- Regional effects (region) are comparable to results from the earlier models. As expected, salmon reach the upstream regions (lower Sacramento and North Delta) before reaching the Central Delta by about 8 days for the Lower Sacramento and by about 2.5 days for the North Delta. Timing at the South Delta cannot be distinguished from the Central Delta.

- Change over time (`wy_meanCentered`; -0.30, CI \[-0.47--0.12\]): There is a credible temporal trend towards earlier migration, although it's modified by the timing of first flush events.

- `sigma` (8.56): Confirms that first flush timing adds modest but real explanatory power.

- Bottom line: First flush timing is a credible predictor of *early* outmigration (10th percentile) but not of bulk or late outmigration, as we hypothesized. If our goal is to understand what drives timing across the entire season, `ff_wyd` is most relevant for characterizing the vanguard of the run.

### priors

Initially, I accepted the default priors that `brms` employs. These weren't bad – 'flat' for all except for the `intercepts` and for `sigma` where it uses `student_t`, but the tuning done for `bmod5` really improved the model outcome.\

```{r}
#| label: priors
#| echo: false

# if you want to see the details...
prior_summary(bmod1)
prior_summary(bmod2)
prior_summary(bmod3)
prior_summary(bmod4)
prior_summary(bmod5)

```

## next steps

Can I look at this for WR only? Include other LAD-based run designations?

Try the mixed effects modeling approach that Jereme recommended.

What happens if I limit the data to genetically-typed salmon but with (necessarily) reduced time series?

Need to consider finer spatial scales: what do accumulation curves for individual sites look like?

What about including the whole year and limiting the time series to post-1992 (or thereabouts)?

Surely there is an applicable metric for the magnitude of a first flush event. Include that in one of these models!

What about other salmon data sources (eg trawls from Chipps Island)? What about other fish species?
