---
title: "salmon outmigrant phenology"
author: "Pete Nelson"
format: 
  html:
    embed-resources: true
execute:
  execute-dir: project
editor: visual
---

# introduction

Beach seine data from the DJFMP dataset holds patterns in the presence and abundance of juvenile Chinook salmon in the lower Sacramento River and the Delta. These data were first collected in 1976 and the most recent data are from water year 2024. The objectives here were to explore the relationship between the timing and magnitude of first flush events, and to examine year-to-year changes in the timing of juvenile Chinook outmigration

# setup

One package *not* available on CRAN: If necessary, get it using `remotes::install_github("doi-usgs/EflowStats@v5.2.0")`

```{r}
#| label: packages
#| message: false

# library(conflicted) # causes WAAAY more trouble than it solves!

library(here)
library(readxl)
library(janitor) 
library(ggplot2)
library(brms)
library(dataRetrieval)
library(EflowStats)
library(fishualize)
library(nlme)
library(Kendall)
library(envPred) # install_github("dbarneche/envPred"), requires devtools
library(tidyverse)

```

# load data

Assumes that you've executed the R code (`edi.244.14.r`) and `dt1:dt5` are in salmon/data/raw.

- United States Fish and Wildlife Service, J. Speegle, C. Macfarlane, E. Huber, D. Goodman, K. Erly, and R. Cook. 2026. Interagency Ecological Program: Over four decades of juvenile fish monitoring data from the San Francisco Estuary, collected by the Delta Juvenile Fish Monitoring Program, 1976-2025 ver 14. Environmental Data Initiative. https://doi.org/10.6073/pasta/ad35c80cafd87ba928d3f37fa124ccaf (Accessed 2026-08-20).

```{r}
#| label: clean-data
#| eval: false

# beach seine data from edi.244.14.5, dt3
# 'raw' includes all spp
raw <- 
  read_rds(here("salmon/data/raw/dt3.rds")) %>% 
  # provide good names
  clean_names() %>% 
  # chng region & station to chr (factors interfere w regularization later)
  # fix date & time variables, originally as chr
  mutate(region = as.character(region_code),
         station = as.character(station_code),
         date = mdy(sample_date),
         time = times(sample_time),
         .keep = "unused") %>% 
  # generate a sample unit (unique seine event) code
  group_by(station, date, time) %>% 
  mutate(id_code = cur_group_id()) %>% 
  relocate(id_code, station, location, date, time) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/raw/raw.rds"))

```

```{r}
#| label: seine-data
#| echo: true

# load clean seine data
raw <-
  read_rds(here("salmon/data/raw/raw.rds"))

# naming conventions
taxonomy <- 
  read_rds(
  here("salmon/data/raw/dt4.rds"))

# site information
sample_sites <- 
  read_rds(
    here("salmon/data/raw/dt5.rds"))


```

## storm data

Let's pull in Rosie's storm data including *first flush* events, both timing and magnitude.

1.  ::: {}
    ```{r}
    #| label: storm-data
    #| message: false
    #| warning: false

    load(here("data/processed/storms/StormData.RData"))
    # Firststorms, Sacflow_wstorms, Storms, StormStartEnd

    # use 'Firststorms'
    storms <-
      Firststorms %>% 
      arrange(WY) %>% 
      mutate(
        migration_year = if_else(month(Date) >= 9, year(Date) + 1, year(Date)),
        ff_wyd = EflowStats::get_waterYearDay(Date, wyMonth = 9L)) %>% 
      select(ID, Date, wy = WY, migration_year, ff_wyd, TotalFlow, SAC, SJR, OUT, YoloSac) %>%
      clean_names() %>% 
      write_rds(here("data/processed/storms/storms.rds"))

    rm(Firststorms, Sacflow_wstorms, Storms, StormStartEnd)

    ```
    :::

# wrangle data

## choose sites

There are beach seine data from a large number of sites, but the consistency with which these sites were sampled varies widely. The code below selects two overlapping sets of sites: those sampled consistently for January–March from water years 1977–2024, and those sampled consistently for November–June from water year 1994–2024. Even at the better sites, there are missing data, particularly for the months April–December prior to 1992, which is why the two windows (1977-2024 & 1994-2024) are treated separately rather than applying one standard to the whole record. For each candidate window, I first required a station to have been in operation for at least 15 years within that window — long enough for a coverage percentage to be meaningful — then required at least 50% of those eligible years to include data from every target month. A handful of stations sitting at or near the threshold were resolved by hand rather than by the rule alone; those exceptions are documented as named overrides in the code.

```{r}
#| label: limit-sites
#| message: false
#| warning: false

# for longest time series (1978-2024)
# list of beach seine stations identified minimally adequate for 
# coverage Jan, Feb & Mar (tibble includes region and region_label)

MIN_YEARS <- 15 # minimum eligible years before the percentage is trusted
THRESHOLD <- 50 # % of the study window with complete target-month coverage
MIN_N_A <- 20 # minimum annual catch (region x wy) before a day10/50/90 benchmark is trusted
MIN_N_B <- 20 # similar distribution to Series A

series_def <- list(
  A = list(months = 1:3, start = 1977, end = 2024), # long time series, Jan-Mar
  B = list(months = c(11, 12, 1:6), start = 1994, end = 2024) # short, Nov-Jun
)

# these sites meet the threshold but they're right at the borderline; I've added
# the manual_keep override because I made the explicit decision to include them
manual_keep <- c("SR094E", "XC001N") # each at threshold (51.1 & 51.0% respectively)

# Lower Sac R station: meets ser A criteria but at 43.8% for ser B (Nov-Jun)
# so this site is retained in A and dropped from B, sites_ltd
keep_SR138E_in_series_B <- FALSE

coverage <- raw %>%
  distinct(station, region, year = year(date), month = month(date)) %>%
  mutate(sampled = 1)

station_coverage_metric <- function(coverage, months, start, end, min_years = MIN_YEARS) {
  n_needed <- length(months)
  
  n_complete <- coverage %>%
    filter(month %in% months, between(year, start, end)) %>%
    count(station, year, name = "n_months") %>%
    filter(n_months == n_needed) %>%
    count(station, name = "n_complete")
  
  first_last <- coverage %>%
    group_by(station) %>%
    summarise(first_year = min(year), last_year = max(year), .groups = "drop")
  
  first_last %>%
    left_join(n_complete, by = "station") %>%
    mutate(
      n_complete   = replace_na(n_complete, 0),
      denom        = pmax(end - pmax(first_year, start) + 1, 0),
      pct_coverage = if_else(denom > 0, round(100 * n_complete / denom, 1), NA_real_),
      eligible     = denom >= min_years
    )
}

# generate a table of stations using first the criteria for series A 
# 1977-2024 but Jan-Mar only
metrics_A <- station_coverage_metric(coverage, series_def$A$months, series_def$A$start, series_def$A$end) %>%
  rename_with(~ paste0(., "_A"), c(n_complete, denom, pct_coverage, eligible))

# criteria for series B (1994-2024, Nov-Jun)
metrics_B <- station_coverage_metric(coverage, series_def$B$months, series_def$B$start, series_def$B$end) %>%
  rename_with(~ paste0(., "_B"), c(n_complete, denom, pct_coverage, eligible))

station_selection <- coverage %>%
  distinct(station, region) %>%
  left_join(metrics_A %>% select(station, first_year, last_year, ends_with("_A")), by = "station") %>%
  left_join(metrics_B %>% select(station, ends_with("_B")), by = "station") %>%
  mutate(
    pass_A = eligible_A & pct_coverage_A >= THRESHOLD,
    pass_B = eligible_B & pct_coverage_B >= THRESHOLD,
    selected_A = pass_A | station %in% manual_keep,
    selected_B = (pass_B | station %in% manual_keep) &
      !(station == "SR138E" & !keep_SR138E_in_series_B)
  )

region_labels <- tibble(
  region = as.character(1:7),
  region_label = factor(c(
    "Lower Sacramento River", "North Delta", "Central Delta", "South Delta",
    "San Joaquin River", "San Francisco & San Pablo Bays", "Sacramento River"
  ))
)

core_delta_regions <- c("1", "2", "3", "4")

station_locations <- sample_sites %>%
  clean_names() %>%
  select(station = station_code, location, latitude, longitude)

# add river kilometer info to the site meta data
station_rkm <- 
  read_rds(here("salmon/data/processed/stations_with_rkm.rds")) %>%
  select(station, river_km) %>%
  mutate(station = as.character(station)) # station_selection uses character; this table has it as a factor

sites <-
  station_selection %>%
  filter(selected_A, region %in% core_delta_regions) %>%
  left_join(region_labels, by = "region") %>%
  left_join(station_locations, by = "station") %>%
  left_join(station_rkm, by = "station") %>%
  mutate(RKI = river_km + 86) %>%
  arrange(region)

sites_ltd <-
  station_selection %>%
  filter(selected_B, region %in% core_delta_regions) %>%
  left_join(region_labels, by = "region") %>%
  left_join(station_locations, by = "station") %>%
  left_join(station_rkm, by = "station") %>%
  mutate(RKI = river_km + 86) %>%
  arrange(region)

sites <- sites %>% mutate(region_label = droplevels(region_label))
sites_ltd <- sites_ltd %>% mutate(region_label = droplevels(region_label))

write_rds(sites, here("salmon/data/processed/sites.rds"))
write_rds(sites_ltd, here("salmon/data/processed/sites_ltd.rds"))

```

## limit seine data

Select seine data to the selected sites or stations and include migration year (Sep 1 - Aug 31) for consistency with the First Flush protocol.

```{r}
#| label: mod-seine
#| echo: true

# filter raw beach seine data to selected stations, using two site series
# dat_A corresponds to Series_A (longer ts, but Jan-Mar only); dat_B corresponds
# to Series_B (shorter ts, but essentially the entire potential migration period)
# adds migration years (day 1 = Sept 1, so a complete year runs Sept 1 - Aug 31)
dat_A <-
  raw %>% 
  filter(station %in% sites$station,
         date >= "1976-09-01" & date < "2024-09-01") %>% 
  mutate(year = year(date),
         month = month(date),
         wy = dataRetrieval::calcWaterYear(date),
         migration_year = if_else(month(date) >= 9, year(date) + 1, year(date)),
         wyd = EflowStats::get_waterYearDay(date, wyMonth = 9L)) %>% 
  relocate(region, .after = id_code) %>% 
  relocate(year, month, wy, migration_year, wyd, .before = date) %>% 
  select(c(station:water_temp, organism_code, fork_length, race_by_length, count, 
                   year, month, wy, migration_year, wyd)) %>% 
  group_by(date, time, station) %>% 
  mutate(id_code = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(id_code) %>% 
  write_rds(., here("salmon/data/raw/dat_A.rds"))

dat_B <-
  raw %>% 
  filter(station %in% sites_ltd$station,
         date >= "1994-09-01" & date < "2024-09-01") %>% 
  mutate(year = year(date),
         month = month(date),
         wy = dataRetrieval::calcWaterYear(date),
         migration_year = if_else(month(date) >= 9, year(date) + 1, year(date)),
         wyd = EflowStats::get_waterYearDay(date, wyMonth = 9L)) %>% 
  relocate(region, .after = id_code) %>% 
  relocate(year, month, wy, migration_year, wyd, .before = date) %>% 
  select(c(station:water_temp, organism_code, fork_length, race_by_length, count, 
                   year, month, wy, migration_year, wyd)) %>% 
  group_by(date, time, station) %>% 
  mutate(id_code = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(id_code) %>% 
  write_rds(., here("salmon/data/raw/dat_B.rds"))

# pivot wide
seine_A <-
  dat_A %>% 
  pivot_wider(
    id_cols = c(id_code:time, water_temp),
    names_from = organism_code,
    values_from = count,
    values_fill = 0,
    values_fn = sum
  ) %>% 
  select(c(id_code:water_temp, CHN)) %>% 
  left_join(sites, by = "station") %>% 
  rename(location = location.x) %>% 
  select(-c(location.y, time)) %>% 
  relocate(c(region, region_label, location, station), .after = id_code) %>% 
  ungroup()

seine_B <-
  dat_B %>% 
  pivot_wider(
    id_cols = c(id_code:time, water_temp),
    names_from = organism_code,
    values_from = count,
    values_fill = 0,
    values_fn = sum
  ) %>% 
  select(c(id_code:water_temp, CHN)) %>% 
  left_join(sites_ltd, by = "station") %>% 
  rename(location = location.x) %>% 
  select(-c(location.y, time)) %>% 
  relocate(c(region, region_label, location, station), .after = id_code) %>% 
  ungroup()

# Series A: Jan-Mar
regional_cumul_A_input <- seine_A %>% 
  filter(month <= 3) %>% 
  group_by(region_label, wy, month, wyd) %>% 
  summarise(daily_total = sum(CHN, na.rm = TRUE), .groups = "drop")

# Series B: Nov-Jun
regional_cumul_B_input <-
  seine_B %>% 
  filter(month %in% c(11, 12, 1:6)) %>% 
  group_by(region_label, wy, month, wyd) %>% 
  summarise(daily_total = sum(CHN, na.rm = TRUE), .groups = "drop")

# pool by region
regional_cumul_A <-
  regional_cumul_A_input %>% 
  group_by(wy, region_label) %>% 
  arrange(region_label, wy, wyd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, month, wyd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/regional_cumul_A.rds"))

regional_cumul_B <-
  regional_cumul_B_input %>% 
  group_by(wy, region_label) %>% 
  arrange(region_label, wy, wyd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, month, wyd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/regional_cumul_B.rds"))

```

## water year types

In California, *water year types* are classifications (Wet, Above Normal, Below Normal, Dry, and Critical) that are used to categorize hydrological conditions over a 12-month period (October 1–September 30), based, for the Sacramento and San Joaquin valleys, on unimpaired runoff. These designations are used for making water management decisions and probably are not the best environmental measures available for understanding salmon ecology, but they are used by managers and offer a reasonable starting point for exploring the relationship between environment and salmon migration.

Add data on *water year types* to the data:

```{r}
#| label: wy-wyd
#| message: false
#| warning: false


# add variables: water year and day of water year
# select data begining with WY 1976 
# (Sep 1 1976 - Aug 1977) and ending with WY 2024 (Sep 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

# NOTE: this unusual choice for water year dates does introduce the possibility of 
# weirdness. 'dataRetrieval::calcWaterYear()' is coded for std federal WY 10/1-9/30
# and can't be modified. fish captured 9/22/08 would be included with WY 2008 (standard)
# instead of WY 2009 (FF protocol). solution: 'migration_year = if_else(month(date) >= 9, year(date) + 1, year(date))' applied in this code chunk and in the storm data chunk



```

# analysis

## accumulation curves

Calculate the accumulated number of salmon over the course of each water year and aggregate over selected beach seine sites. Determine a cumulative percentage for each day over the entire year and for the 3-month period January, February, and March. Plot annual accumulation curves to show the pattern of juvenile outmigration for each water year, as well as for each January-March period for each year. The truncated data are used as these months were sampled consistently throughout the time series; sampling effort during the remaining months, April-December was limited prior to 1992.

```{r}
#| label: counts
#| message: false
#| warning: false

daily_regional_A <-
  seine_A %>% 
  group_by(region_label, wy, month, wyd) %>% 
  summarise(daily_total = sum(CHN, na.rm = TRUE), .groups = "drop")

daily_regional_B <-
  seine_B %>% 
  group_by(region_label, wy, month, wyd) %>% 
  summarise(daily_total = sum(CHN, na.rm = TRUE), .groups = "drop")

```

Plot the accumulation curves.

```{r}
#| label: annual-accumulation
#| message: false
#| warning: false

# accumulation curves at the regional level
# Series A, Jan-Mar data, 1978-2024
daily_regional_A %>% 
  filter(month <= 3) |> 
  group_by(region_label, wy) %>% 
  arrange(wy, wyd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, wyd) %>% 
  ggplot(aes(x = wyd, y = cumul_percent, group = interaction(region_label, wy), color = wy)) +         
  geom_line() +
  coord_cartesian(xlim = c(120, 220)) +
  geom_hline(yintercept = c(10, 50, 90), linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year", y = "Cumulative % of 3-Month Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "Series A; sites summed by region; Jan-Mar data from 1978-2024",
       color = "water year") +
  theme_bw()

# Series B, Nov-Jun data, 1994-2024
daily_regional_B %>% 
  filter(month %in% c(11, 12, 1:6)) %>% 
  group_by(region_label, wy) %>% 
  arrange(wy, wyd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, wyd) %>% 
  ggplot(aes(x = wyd, y = cumul_percent, group = interaction(region_label, wy), color = wy)) +         
  geom_line() +
  coord_cartesian(xlim = c(60, 305)) +  # just the visual zoom window, doesn't need to be exact
  geom_hline(yintercept = c(10, 50, 90), linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year", y = "Cumulative % of 8-Month Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "Series B; sites summed by region; Nov-Jun data from 1994-2024",
       color = "water year") +
  theme_bw()

```

```{r}
#| label: 3month-accumulation
#| message: false
#| warning: false

# Series A
regional_cumul_A %>% 
  arrange(region_label, wy, wyd) %>%
  mutate(region_label = 
           fct_relevel(region_label, 
                       "Lower Sacramento River", "North Delta", 
                       "Central Delta", "South Delta")) %>% 
  ggplot(aes(x = wyd,
             y = cumul_percent,
             color = wy,
             group = wy)) +
  geom_line() +
  xlim(120, 220) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of 3-Month Total", 
       title = "Juvenile Chinook Regional Accumulation Curves",
       subtitle = "Series A; January - March, 1977-2024",
       color = "Water Year") +
  facet_wrap(~region_label, nrow = 2) +
  theme_bw()

# Series B
regional_cumul_B %>% 
  arrange(region_label, wy, wyd) %>%
  mutate(region_label = 
           fct_relevel(region_label, 
                       "Lower Sacramento River", "North Delta", 
                       "Central Delta", "South Delta")) %>% 
  ggplot(aes(x = wyd,
             y = cumul_percent,
             color = wy,
             group = wy)) +
  geom_line() +
  xlim(60, 305) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of 8-Month Total", 
       title = "Juvenile Chinook Regional Accumulation Curves",
       subtitle = "Series B; November - June, 1994-2024",
       color = "Water Year") +
  facet_wrap(~region_label, nrow = 2) +
  theme_bw()

```

## benchmarks

Determine the day of each water year when 10%, 50% and 90% of the total annual numbers are reached.

```{r}
#| label: benchmarks
#| message: false

## ID benchmarks ----
# day of water year when the 10%, 50% and 90% of total annual numbers--accumulation benchmarks

three_month_regional_benchmarks_A <-
  regional_cumul_A %>% 
  arrange(region_label, wy, wyd) %>% 
  group_by(region_label, wy) %>% 
  summarise(
    annual_total = first(annual_total),
    day_10 = min(wyd[cumul_percent >= 10]),
    day_50 = min(wyd[cumul_percent >= 50]),
    day_90 = min(wyd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  filter(annual_total >= MIN_N_A) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), names_to = "percentile", values_to = "wyd") %>% 
  mutate(percentile = factor(percentile),
         wy_meanCentered = wy - mean(wy))

three_month_regional_benchmarks_B <-
  regional_cumul_B %>% 
  arrange(region_label, wy, wyd) %>% 
  group_by(region_label, wy) %>% 
  summarise(
    annual_total = first(annual_total),
    day_10 = min(wyd[cumul_percent >= 10]),
    day_50 = min(wyd[cumul_percent >= 50]),
    day_90 = min(wyd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  filter(annual_total >= MIN_N_B) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), names_to = "percentile", values_to = "wyd") %>% 
  mutate(percentile = factor(percentile),
         wy_meanCentered = wy - mean(wy))



```

## water year type

Add the water year type data (loaded previously) to the percentile data for analysis and graphics.

```{r}
#| label: wy-type
#| message: false
#| warning: false

## year type ----
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(str_trim(WYT)),
         watershed = factor(Area),
         .keep = "unused")

three_month_regional_benchmarks_A <-
  wy_type %>% 
  filter(watershed == "Sacramento Valley") %>% 
  right_join(three_month_regional_benchmarks_A, by = "wy") %>% 
  select(-c(watershed)) %>% 
  # make wyt an ordered factor
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      ordered = TRUE)) %>% 
  relocate(wy_meanCentered, .after = wy) %>% 
  write_rds(here("salmon/data/processed/three_month_regional_benchmarks_A.rds"))

three_month_regional_benchmarks_B <-
  wy_type %>% 
  filter(watershed == "Sacramento Valley") %>% 
  right_join(three_month_regional_benchmarks_B, by = "wy") %>% 
  select(-c(watershed)) %>% 
  # make wyt an ordered factor
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      ordered = TRUE)) %>% 
  relocate(wy_meanCentered, .after = wy) %>% 
  write_rds(here("salmon/data/processed/three_month_regional_benchmarks_B.rds"))

```

## migration timing

Plot the water year days when each benchmark (10, 50, or 90%) is reach for each water year.

```{r}
#| label: wy-effect
#| message: false
#| warning: false

# begin by writing a function to plot benchmark timing
plot_benchmark_timing <- function(data, pct, series_label, window_label) {
  pct_num <- str_remove(pct, "day_")  # "day_10" -> "10", just for labeling
  
  data %>%
    filter(percentile == pct) %>%
    mutate(
      region_label = factor(
        region_label,
        levels = c("Lower Sacramento River", "North Delta", "Central Delta", "South Delta")
      )
    ) %>%
    ggplot(aes(x = wy, y = wyd, color = region_label)) +
    geom_point(size = 4) +
    scale_color_fish(discrete = TRUE, option = "Oncorhynchus_tshawytscha", alpha = 0.6) +
    geom_smooth(alpha = 0.2, method = "lm", se = TRUE) +
    theme_bw() +
    theme(legend.position = "none") +
    facet_wrap(~ region_label, ncol = 2) +
    labs(title = paste0("Chinook Salmon Outmigration Timing, ", pct_num, "% Benchmark"),
         subtitle = paste0("Series ", series_label, "; data from ", window_label, " only"),
         x = "Water Year",
         y = paste0(pct_num, "% of Total Reached (day of water year)"))
}

# now call the function once for each percentile, for each series

percentiles <- c("day_10", "day_50", "day_90")

# quick QA check, same as before, once per series
unique(three_month_regional_benchmarks_A$region_label)
unique(three_month_regional_benchmarks_B$region_label)

walk(percentiles, ~{
  p <- plot_benchmark_timing(three_month_regional_benchmarks_A, .x, "A", "January - March")
  print(p)
  ggsave(here("salmon/plots", paste0("benchmark_", str_remove(.x, "day_"), "pct_A.png")), plot = p)
})

walk(percentiles, ~{
  p <- plot_benchmark_timing(three_month_regional_benchmarks_B, .x, "B", "November - June")
  print(p)
  ggsave(here("salmon/plots", paste0("benchmark_", str_remove(.x, "day_"), "pct_B.png")), plot = p)
})


```

## models

### change over time and water year type effects

Testing for potential effects of water year (i.e., change over time), water year type, regional differences, and differences in the three benchmarks (is there a difference in the slopes).

```{r}
#| label: bayes1
#| message: false
#| warning: false

# ensure that we've got a nested directory
# output_dir <- here("salmon", "results")
# if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# I've used a Gaussian distribution (at least for now) because the distributions 
# for each benchmark look "normalish." See
three_month_regional_benchmarks %>%
  ggplot(aes(wyd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")
# I'll do a pp_check of the model; if this shows the model giving wyd=60 or wyd=250 with
# non-trivial probabilities, I'll have to swtich to Beta or truncated normal distribution

bmod1 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = gaussian,
      wyd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      # set seed
      seed = 42,
      # brms automatically appends '.rds'
      file = here("salmon/results/bmod1"),
      file_refit = "on_change"
      )

pp_check(bmod1) # looks good!
(bmod1_sum <- summary(bmod1)) # see analysis
plot(bmod1) # looks good

conditional_effects(bmod1, "wy_meanCentered:percentile") # plot doesn't add much

```

All `Rhat = 1.00`, so the model converged cleanly. We have effective sample sizes in the low thousands and `n = 507`, so we can take the estimates seriously.

Without going into the output details here (I will with the best model), the entire distribution of the `wyd` benchmarks (10th, 50th, 90th percentiles) is moving more than 0.4 days earlier each year, and the shift is statistically credible. The three benchmarks are shifting in parallel rather than diverging or converging. Fish arrive in the Lower Sacramento River about a week prior to their arrival in the Delta, and timing to reach sites within the Delta are indistinguishable from each other. Arrival times are slightly earlier during wet years than in dry years, but that gap has narrowed over time. However, water year hydrology is genuinely heavy-tailed — ENSO events and atmospheric rivers can produce years that are extreme outliers relative to the typical range (see the accumulation curves for 1977 above for example), and I don't want those years unduly affecting my coefficient estimates. A Gaussian model will try harder to accommodate outliers by distorting the fitted mean, whereas a Students t-distribution is more robust to these extreme values and less likely to overfit the model.

I ran the same model using a Students t-distribution (`bmod2`, below):

```{r}
#| label: robust
#| message: false
#| warning: false

bmod2 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = student,
      wyd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod2"),
      file_refit = "on_change"
      )

pp_check(bmod2) # looks good!
(bmod2_sum <- summary(bmod2)) # see analysis
plot(bmod2) # looks good

conditional_effects(bmod2, "wy_meanCentered:percentile") # doesn't add much

```

This a real improvement over `bmod1` with the `nu` estimate \<10. I compare these two models directly later in this report, but first let's start looking at the possible effects of first flush events.

### first flush effect

Pull in Rosie's storm data and integrate that with the benchmark data from above.

```{r}
#| label: ff-data
#| message: false
#| warning: false

# no first flush storms for 1972, 1976, 1977, 2020, or 2021 (NA wyd -> dropped below)

ff_data_A <- 
  three_month_regional_benchmarks_A %>% 
  left_join(storms, 
            by = "wy",
            keep = FALSE) %>% 
  mutate(region = region_label,
         .keep = "unused") %>% 
  filter(!is.na(ff_wyd)) %>% 
  select(-c(id, date)) %>% 
  relocate(c(region, wyd, ff_wyd, total_flow), .after = wyt)

ff_data_B <- 
  three_month_regional_benchmarks_B %>% 
  left_join(storms, 
            by = "wy",
            keep = FALSE) %>% 
  mutate(region = region_label,
         .keep = "unused") %>% 
  filter(!is.na(ff_wyd)) %>% 
  select(-c(id, date)) %>% 
  relocate(c(region, wyd, ff_wyd, total_flow), .after = wyt)

```

### first flush timing

The migration data look cool, but first, does the timing of first flush events change over time? Plotting the timing and then a simple linear regression suggests that first flush events have come later over this time period.

```{r}
#| label: ff-timing

# does the timing of the ff events change over time?
ff_data_A %>% 
  filter(percentile == "day_10", region == "Central Delta") %>% 
  ggplot(aes(x = wy, y = ff_wyd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Day of First Flush",
       subtitle = "Series A; 1978 - 2024",
       x = "Water Year",
       y = "Day of First Flush (day of water year)") +
  theme_bw()

lm_ff_annual <-
  lm(data = ff_data_A %>% 
       # ensure we aren't pseudo-replicating
       filter(percentile == "day_10", region == "Central Delta"), 
     ff_wyd ~ wy)

summary(lm_ff_annual) # no. there's a slope but way noisy

# evidence of autocorrelation?
acf(residuals(lm_ff_annual))
# nope

# what's the potential role of water year type here, given that it's almost
# certainly correlated with the timing of first flush events?
lm_ff_wyt <- lm(
  data = ff_data_A %>% 
    filter(percentile == "day_10", region == "Central Delta"),
  ff_wyd ~ wy + wyt
)
summary(lm_ff_wyt)

# let's compare these models
AIC(lm_ff_annual, lm_ff_wyt)
anova(lm_ff_annual, lm_ff_wyt)

ff_data_B %>% 
  filter(percentile == "day_10", 
         region == "Central Delta") %>% 
  ggplot(aes(x = wy, y = ff_wyd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Day of First Flush",
       subtitle = "Series B; 1998 - 2024",
       x = "Water Year",
       y = "Day of First Flush (day of water year)") +
  theme_bw()

# is there a trend? can't think of a climate- or a management-related reason
MannKendall(storms$ff_wyd)
MannKendall(storms$total_flow)

```

The Mann-Kendall test shows no statistically significant monotonic trend in first flush timing (tau = 0.0958, p = 0.18) or magnitude over time (tau = 0.09, p = 0.17).

Let's see if there are any patterns by plotting some of these data.

```{r}
#| label: plots
#| message: false
#| warning: false


ff_data %>%
  ggplot(aes(wyd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")
# looks good; skew doesn't appear excessive

ff_data %>% 
  ggplot(aes(x = ff_wyd, y = wyd, color = percentile)) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(title = "First Flush and Chinook Outmigration Timing",
       subtitle = "outmigration during Jan - Mar, 1978-2024, only",
       x = "First Flush Timing (day of water year)",
       y = "% Accumulation Benchmark (day of water year)") +
  theme_bw()

ggsave(here("salmon/plots/first_flush_timing.png"))
```

The plot suggested that *maybe* these events were happening later over time, but the regression analysis shows that, while there is an effect of water year *type* on the timing of first flush, there's insufficient evidence to suggest that there's been a change over time. The improvement to the model with the addition of `wyt` is modest (p=0.056), but seems real to me.

### first flush effects on migration timing

Now the question is, does the timing of first flush events affect the timing of juvenile salmon outmigration, as the figure relating migration timing to the timing of first flush events above suggests? I expect that *when* we get that first flush will affect the timing for that 10% benchmark, but that the effect will be reduced or absent with regard to reaching the 50 and 90% benchmarks. Note that we don't have juvenile salmon accumulation data from South Delta for years 1989, 1991, 1993, 1997, 2009, 2010, 2012-2016, 2018, 2019, 2022, and 2023.

```{r}
#| label: ff-models
#| message: false
#| warning: false

bmod3 <- 
  brm(data = ff_data,
      # 1st interaction lets slope of wyd differ across benchmarks
      # I expect that day_10 will track ff closely but day_90 not so much
      wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered,
      # student-t helped in bmod2, so
      family = student,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod3"),
      file_refit = "on_change")

pp_check(bmod3) # looks good!
(bmod3_sum <- summary(bmod3)) # nu=7.21; student distribution is helping
plot(bmod3) # looks good
conditional_effects(bmod3, "ff_wyd:percentile") # ff matters more early in migr

bmod4 <- 
  brm(data = ff_data,
      # within yr obs may not be independent; this may help
      wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy),
      family = student,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod4"),
      file_refit = "on_change")

pp_check(bmod4) # looks good!
(bmod4_sum <- summary(bmod4)) # regional effects emerge more clearly
plot(bmod4) # looks good
conditional_effects(bmod4, "ff_wyd:percentile") # dividing the migration into
# early, mid and late allows us to see there's a real effect of the ff timing
# albeit primarily on those first, early migrants

# define some priors based earlier model output
my_priors <- c(
  prior(normal(150, 50), class = Intercept),
  prior(normal(0, 1), class = b, coef = ff_wyd),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_50"),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_90"),
  prior(normal(0, 2), class = b, coef = wy_meanCentered),
  prior(normal(0, 30), class = b), # now only hits percentile & region contrasts
  prior(normal(0, 20), class = sd),
  prior(exponential(1), class = sigma),
  prior(gamma(2, 0.1), class = nu)
)

bmod5 <- brm(
  data = ff_data,
  wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy),
  family = student,
  prior = my_priors,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 42,
  file = here("salmon/results/bmod5"),
  file_refit = "on_change"
)

pp_check(bmod5) # looks GREAT
pp_check(bmod5, type = "stat", stat = "mean")
(bmod5_sum <- summary(bmod5)) # what's going on w S Delta? SJR fish?!
plot(bmod5) # looks good
conditional_effects(bmod5, "ff_wyd:percentile")

```

The diagnostics look clean again — all `Rhat` = 1.00, ESS well above 1000, and `nu` well below 10. The results look really good, but I think we should add water year type to the model to see if overall water year conditions affect migration timing. This could, for example, affect the tail end of the migration (`day_90`) while having little or no effect on the first 10%.

```{r}
#| label: ff+wyt

# we're going to have to revise the priors though:
get_prior(
  wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + mo(wyt) + (1 | wy),
  data = ff_data,
  family = student
) # this is mostly to ensure that I'm using the right terms

my_priors_bmod6 <- c(
  prior(normal(150, 50), class = Intercept),
  prior(normal(0, 1), class = b, coef = ff_wyd),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_50"),
  prior(normal(0, 1), class = b, coef = "ff_wyd:percentileday_90"),
  prior(normal(0, 2), class = b, coef = wy_meanCentered),
  prior(normal(0, 30), class = b, coef = mowyt),
  prior(normal(0, 30), class = b), # percentile & region contrasts
  prior(normal(0, 20), class = sd),
  prior(exponential(1), class = sigma),
  prior(gamma(2, 0.1), class = nu),
  prior(dirichlet(1), class = simo, coef = mowyt1)
)

bmod6 <- brm(
  data = ff_data,
  # note addition of the monotonic wyt
  wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + mo(wyt) + (1 | wy),
  # sample_prior = "only", # run pp_check(bmod6); if ok run without
  family = student,
  prior = my_priors_bmod6,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 42,
  file = here("salmon/results/bmod6"),
  file_refit = "on_change"
)

pp_check(bmod6)
pp_check(bmod6, type = "stat", stat = "mean")
(bmod6_sum <- summary(bmod6)) # what's going on w S Delta? SJR fish?!
plot(bmod6) # looks good
conditional_effects(bmod6, "ff_wyd:percentile")

# does ff magnitude ('yolo_sac') have an effect?
# first, is there correlation btwn ff_wyd and yolo_sac? seems unlikely but-
cor(ff_data$ff_wyd, ff_data$yolo_sac, use = "complete.obs")
# nope!

bmod7 <- brm(
  data = ff_data,
  # note DROPPING of the monotonic wyt & addition of yolo_sac (ff mag)
  wyd ~ 1 + ff_wyd * percentile + ff_wyd * yolo_sac + percentile * yolo_sac + 
    region + wy_meanCentered + (1 | wy),
  # sample_prior = "only", # run pp_check(bmod6); if ok run without
  family = student,
  # prior = my_priors_bmod6,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 42,
  file = here("salmon/results/bmod7"),
  file_refit = "on_change"
)

pp_check(bmod7)
pp_check(bmod7, type = "stat", stat = "mean")
(bmod7_sum <- summary(bmod7)) # model prob over-parameterized
plot(bmod7)
conditional_effects(bmod7, "ff_wyd:percentile")

bmod8 <- brm(
  data = ff_data,
  wyd ~ 1 + ff_wyd * percentile + yolo_sac * percentile +
    region + wy_meanCentered + (1 | wy),
  family = student,
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  # improve sampling efficiency 
  control = list(max_treedepth = 15),
  seed = 42,
  file = here("salmon/results/bmod8"),
  file_refit = "on_change"
)
pp_check(bmod8) # looks okay
pp_check(bmod8, type = "stat", stat = "mean")
plot(bmod8)
(bmod8_sum <- summary(bmod8))
# check specific parameter pairs you care about
pairs(bmod8, pars = c("b_ff_wyd", "b_yolo_sac", 
                      "b_percentileday_50:yolo_sac",
                      "b_percentileday_90:yolo_sac"))

bmod5 <- add_criterion(bmod5, "loo")
bmod6 <- add_criterion(bmod6, "loo")
# bmod7 <- add_criterion(bmod7, "loo") # too messed up to use
bmod8 <- add_criterion(bmod8, "loo")
loo_compare(bmod5, bmod6, bmod8)

```

The model including water year type (`bmod6`) showed negligible improvement in predictive accuracy over the simpler model, `bmod5`, (Δelpd = –0.3 ± 0.2), and the water year type coefficient was not credibly different from zero, so we should go with the more parsimonious `bmod5`.

Here's a detailed review of the bmod5 output:

- `Intercept` (101.26, CI \[93.36-109.29\]): This is the expected `wyd` for the 10th percentile benchmark in the reference region (Central Delta) at the mean year when `ff_wyd` = 0. Since day 0 of the water year isn't meaningful biologically, treat this more as a mathematical anchor than an interpretable quantity.

- Day of the first flush (`ff_wyd`; 0.21, CI \[0.13-0.29\]): Credibly positive — each day later that first flush occurs, the 10th percentile outmigration benchmark shifts \~0.21 days later. First flush timing matters: a pulse that comes a month late pushes early migrants roughly 6 days later.

- Interactions between first flush timing and the benchmarks:

  - 10th percentile: **+0.21** (the main effect)

  - 50th percentile: 0.21 – 0.15 = **+0.06** (barely a relationship)

  - 90th percentile: 0.21 – 0.23 = **-0.02** (negligible)

- The timing of first flush events clearly affects when the first juvenile migrants reach the Delta. Presumably other factors (water temperature, cumulative flow conditions, individual variation) determine when the bulk of the outmigrants get there.

- Regional effects (`region`) are comparable to results from the first models. The lower Sacramento is about 8 days earlier than the reference region; fish reach the North Delta a little more than 2 days before the Central Delta. Timing for the South Delta can't be distinguished from the reference region. The latter result could be due to missing data for a large number of years, the effect of San Joaquin river fish entering the Delta independently, or both.

- Change over time (`wy_meanCentered`; -0.30, CI \[-0.47--0.12\]): There is a credible temporal trend towards earlier migration.

- sigma (8.56): Smaller than `bmod1` (14.62); confirms that first flush timing adds real explanatory power.

- Bottom line: There has been a shift in the timing of salmon outmigration over this 45 year period, with the first 10% of each year's cohort now reaching the Delta nearly two weeks earlier than the late 1970's. First flush timing is also a credible predictor of *early* outmigration (10th percentile) but not of bulk or late outmigration, as we hypothesized. Outmigration phenology appears to be changing over the long term (decades), but is also strongly affected by year-to-year differences.

Compare model performance.

```{r}
#| label: pareto-k

models <- 
  list(bmod1 = bmod1, bmod2 = bmod2, bmod3 = bmod3, bmod4 = bmod4, bmod5 = bmod5, bmod6 = bmod6)

# fit loo for each
loos <- map(models, loo)

# summary table of k diagnostics per model
k_summary <- imap_dfr(loos, ~{
  k <- .x$diagnostics$pareto_k
  tibble(
    model     = .y,
    n_obs     = length(k),
    k_max     = max(k),
    n_ok      = sum(k <= 0.5),
    n_okish   = sum(k > 0.5 & k <= 0.7),
    n_bad     = sum(k > 0.7 & k <= 1.0),
    n_veryBad = sum(k > 1.0)
  )
})
print(k_summary)

# detailed table of every observation with k > 0.7
bad_points <- imap_dfr(loos, ~{
  k <- .x$diagnostics$pareto_k
  idx <- which(k > 0.7)
  tibble(model = .y, obs = idx, pareto_k = k[idx])
}) |>
  arrange(desc(pareto_k))

if (nrow(bad_points) == 0) {
  message("All Pareto k values <= 0.7 across all models.")
} else {
  message("Observations with Pareto k > 0.7:")
  print(bad_points)
}

```

```{r}
#| label: loo


# loo_compare(loos$bmod1, loos$bmod2, loos$bmod3, loos$bmod4)
# can't compare all four directly bc unequal sample sizes

loo_compare(loos[c("bmod1", "bmod2")])
loo_compare(loos[c("bmod3", "bmod4", "bmod5", "bmod6")])

```

Clean result — bmod2 (Student-t) is clearly preferred over bmod1 (Gaussian), with p_worse=0.94 meaning there's a 94% probability bmod1 is worse. This makes sense given that Student-t is more robust to outliers and our wyd data likely has some extreme values in years with unusual migration timing. Models `bmod2` and `bmod5` outperform their respective alternatives.

So bmod4 and bmod5 use identical formulas and data, differing only in priors — bmod4 uses brms defaults, bmod5 uses weakly informative priors. The LOO comparison showing them as virtually identical (elpd_diff = -0.1) means the priors have negligible influence on the posterior, which is actually reassuring — it suggests the data are informative enough to dominate regardless of prior choice. Given that, **bmod4 (default priors) is the simpler and equally well-supported choice**. The weakly informative priors in bmod5 aren't hurting anything but they're not helping either, so there's no strong justification for the added complexity of specifying them.

There are multiple lines of evidence supporting bmod4 over bmod6:

1.  **LOO:** elpd_diff = -0.2, essentially noise — wyt adds nothing predictively

2.  **Posterior for mowyt:** Estimate = -0.09, 95% CI \[-1.63, 1.55\] — completely straddles zero with wide uncertainty, no credible effect

3.  **Parsimony:** `(1|wy)` already captures year-to-year variation in a flexible, data-driven way. `mo(wyt)` imposes an ordinal structure on top of that which the data don't support

4.  **Conceptual:** ff_wyd is already carrying the "what kind of year is this" signal — wyt and ff_wyd are likely correlated, so wyt has little independent explanatory power left once ff_wyd is in the model. The combination of no predictive improvement, negligible and uncertain coefficient, and conceptual redundancy with ff_wyd and `(1|wy)` makes a strong case. bmod4 is your model.

### interpretation

While output from the two models designed to study the importance of first flush timing on migration phenology are similar, `bmod5` outperformed `bmod3` and `bmod4` so let's work through the output for `bmod4`. The diagnostics look clean again — all Rhat = 1.00, ESS well above 1000. How should we understand these results?

- `Intercept` (101.26, CI \[93.36-109.29\]): This is the expected `wyd` for the 10th percentile benchmark in the reference region (Central Delta) at the mean year when `ff_wyd` = 0. Since day 0 of the water year isn't meaningful biologically, treat this more as a mathematical anchor than an interpretable quantity.

- Day of the first flush (`ff_wyd`; 0.21, CI \[0.13-0.29\]): Credibly positive — each day later that first flush occurs, the 10th percentile outmigration benchmark shifts \~0.21 days later. First flush timing matters: a pulse that comes a month late pushes early migrants about 6 days later.

- Interactions between first flush timing and the benchmarks:

  - 10th percentile: **+0.21** (the main effect)

  - 50th percentile: 0.21 – 0.15 = **+0.06** (barely a relationship)

  - 90th percentile: 0.21 – -0.23 = **-0.02** (even less of a relationship)

- The timing of first flush (`ff_wyd`) events clearly impacts when the first juvenile migrants (`percentileday_10`) reach the Delta. Presumably other factors (water temperature, cumulative flow conditions, individual variation) determine when the bulk of the outmigrants get there (`percentileday_50` and `percentileday_90`).

- Regional effects (region) are comparable to results from the earlier models. As expected, salmon reach the upstream regions (lower Sacramento and North Delta) before reaching the Central Delta by about 8 days for the Lower Sacramento and by about 2.5 days for the North Delta. Timing at the South Delta cannot be distinguished from the Central Delta.

- Change over time (`wy_meanCentered`; -0.30, CI \[-0.47--0.12\]): There is a credible temporal trend towards earlier migration, although it's modified by the timing of first flush events.

- `sigma` (8.56): Confirms that first flush timing adds modest but real explanatory power.

- Bottom line: First flush timing is a credible predictor of *early* outmigration (10th percentile) but not of bulk or late outmigration, as we hypothesized. If our goal is to understand what drives timing across the entire season, `ff_wyd` is most relevant for characterizing the vanguard of the run.

### priors

Initially, I accepted the default priors that `brms` employs. These weren't bad – 'flat' for all except for the `intercepts` and for `sigma` where it uses `student_t`, but the tuning done for `bmod5` really improved the model outcome.\

```{r}
#| label: priors
#| echo: false

# if you want to see the details
prior_summary(bmod1)
prior_summary(bmod2)
prior_summary(bmod3)
prior_summary(bmod4)
prior_summary(bmod5)

```

Scrap `bmod7` — the Rhat values are catastrophic (3.1–3.8, should be ≤1.01). The chains never converged, Bulk_ESS of 4-5 (should be \>\>100), and that intercept estimate of -27,000 with a massive SE tells us that the model is completely unidentified. This isn't a biological signal problem, it's a severe collinearity/identifiability problem.

I checked for a correlation between the timing of the first flush (`ff_wyd`) and the magnitude of that event (`yolo_sac`) using `cor(ff_data$ff_wyd, ff_data$yolo_sac, use = "complete.obs")` but there's no correlation (result = -0.0602). For this reason, I dropped `ff_wyd * yolo_sac` from the model and ran `bmod8` (also increasing the treedepth, max = 15). These results essentially confirm that magnitude adds nothing to the model and that we should stick with `bmod4` as our working model.

## first flush conclusions

The model contenders are:

- **bmod4**: default priors, no `wyt` (water year type)

- **bmod5**: weakly informative priors, no wyt

- **bmod6**: default priors, with `wyt`

The case for **bmod4** as the winner:

- `bmod1/2` aren't comparable (different n) but `bmod2` \> `bmod1` confirms Student-t over Gaussian.

- vs `bmod5`: LOO essentially identical, priors don't matter, default priors are simpler and more transparent — **`bmod4` wins**

- vs `bmod6`: LOO essentially identical, wyt coefficient negligible and uncertain, `(1|wy)` already captures year-type variation — **`bmod4` wins**

- `bmod4`, `bmod5`, `bmod6`, and `bmod8` all have virtually identical `elpd_loo` (-1878.0 to -1878.3) — within noise of each other. Adding `wyt` (`bmod6`), weakly informative priors (`bmod5`), or `yolo_sac` (`bmod8`) buys nothing.

- `bmod3` (no random effect) is substantially worse (`elpd_loo` -1932 vs -1878) — confirms `(1|wy)` is essential.

- `bmod7` is a disaster (`elpd_loo` -11073, `p_loo` of 7322 meaning the model is wildly overfit, 47 bad Pareto k values) — confirms the three-way interaction was completely unidentified.

Our model is `bmod4`:

`wyd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy)`

The decision is defensible on three grounds: predictive performance (LOO), parsimony, and the biological argument that `ff_wyd + (1|wy) + wy_meanCentered` already capture what `wyt` would add. The `p_loo` for `bmod4` (\~49) vs `bmod3` (\~14) is worth noting — the random effects are consuming \~35 effective parameters, which is expected with 45 year levels and confirms the model is using that structure meaningfully rather than overfitting.

### model output specifics

We're explaining `wyd` (day of the water year when benchmarks are reached--10, 50 and 90% of the cohort) as a function of the interaction between first flush timing (`ff_wyd`) and which benchmark we're meeting (`percentile`) plus the `region` of the Delta we're sampling plus the year (`wy_meanCentered`, because years are expected to differ for a variety of potential reasons) plus the water year intercept (`(1 | wy)`). We used the student distribution (vs gaussian or other) because of heavy tails--outliers at both extremes (`nu`=4.3). All `Rhat` = 1; the model converged cleanly. The variance around the water year intercept (`sd(Intercept)`) is roughly 7 days; this tells us how much average benchmark timing varies across water years after accounting for predictors. `sigma` (\~8.85 days) is the residual observation-level noise — unexplained variation within years across regions and benchmarks.

For every day the first flush event arrives later, we expect migration timing to be pushed later about 0.22 days; credible intervals are reasonably tight. Later benchmarks are reached later--no surprise; this is how they're calculated! Fish reach Lower Sacramento River and North Delta before they reach the Central Delta by about 8 and 2.5 days respectively. Timing at the South Delta isn't substantially different from arrival at the Central Delta. There's a trend with respect to year (`wy_meanCentered`): over the course of this study, for every year, migration generally happens sooner by about 0.3 days, nearly 2 weeks over the 46 year time series. The main effect of `ff_wyd` (0.22) applies to the d10 benchmark (the reference level). The interaction terms modify that slope for the other benchmarks:

- d50 slope = 0.22 + (-0.16) = 0.06: later first flush barely affects when 50% of fish have passed

- d90 slope = 0.22 + (-0.24) = -0.02: essentially zero effect of first flush timing on d90

Thus, **first flush timing primarily affects early migrants**. When the first flush arrives late, the leading edge of migration is pushed later, but the bulk and tail of the migration are largely unaffected — they arrive on a more fixed schedule regardless of when the first flush occurred. Effective sample size (ESS) values (1200-6000) are well above any threshold of concern indicating that our posterior summaries are stable.

## to do

# (working with `thresholds_annual`)

- center water year (wyd) DONE

- add FF dates DONE

- add wyt?

- calculate env pred for each...region? station?? using d10, d50, d90??? DONE

- station should probably be a factor

- run analyses

## winter run

What LAD winter run show up in the beach seine data?

```{r}
#| label: wr-plots

# what size distribution (under 150mm FL) do we see in winter run?
raw |> 
  filter(organism_code == "CHN" & fork_length < 150 & fork_length > 0) |>
  ggplot(aes(x = fork_length)) +
  geom_histogram(bins = 30, color = "white", fill = "salmon") +
  labs(title = "Beach Seine Chinook < 150mm FL",
       x = "Fork Length (mm)") +
  theme_bw()

# select winter run (<150mmTL) beach seine counts
wr <-
  dat %>% 
  filter(race_by_length == "Winter" & fork_length < 150 & fork_length > 0) |> 
  pivot_wider(
    # relies on 'id_code' to distinguish among sample units, but retains environmental data
    id_cols = c(id_code:time, water_temp),
    # columns for all fish spp (only CHN needed for this analysis)
    names_from = organism_code,
    # log number of each speciesxfork_length for each seine (id_code)
    values_from = count,
    # if no matches, enter '0'
    values_fill = 0,
    # sum counts if multiple matches
    values_fn = sum
  ) %>% 
  select(c(id_code:water_temp, winter_run = CHN)) %>% 
  left_join(sites, by = "station") %>% 
  rename(location = location.x) %>% 
  select(-c(location.y, year, month, time)) %>% 
  relocate(c(region, region_label, location, station), 
           .after = id_code) %>% 
  ungroup()

# daily wr catches aggregated by region
daily_wr_regional <-
  wr %>% 
  group_by(region_label, wy, month = month(date), wyd) %>% 
  summarise(
    daily_total = sum(winter_run, na.rm = TRUE),
    .groups = "drop"
  )

# what months are LAD winter run caught?
daily_wr_regional |> 
  ggplot(aes(x = as.factor(month))) +
  geom_bar() +
  ylim(0, 390) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Beach Seine LAD Winter Run Chinook",
       subtitle = "< 150mm FL",
       x = "Month") +
  theme_bw()

```

Develop some winter run accumulation curves.

```{r}
#| label: wr-accumul

# numbers were really low even Delta-wide (code not shown)
daily_wr_regional %>%
  group_by(wy) %>%
  summarise(total = sum(daily_total)) %>%
  arrange(wy) |> 
  ggplot(aes(x = wy, y = total)) +
  labs(title = "Beach Seine LAD Winter Run",
       x = "Water Year",
       y = "Total Count") +
  geom_col() +
  theme_bw()

# accumulation curves
daily_wr_regional %>% 
  filter(wy >= 1998) %>% 
  group_by(wy) %>% 
  arrange(wy, wyd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, wyd) %>% 
  ggplot(aes(x = wyd,
             y = cumul_percent,
             # identify unique station*wy combos
             # group = interaction(station, wy),
             color = wy)) +         
  geom_line() +
  # keep the data but zoom the view
  coord_cartesian(xlim = c(0, 300)) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile LAD WR Chinook Accumulation Curves",
       subtitle = "Delta-wide; data from 1998-2024",
       color = "water year") +
  theme_bw()

```

Clearly, there simply are not enough winter run juveniles to pursue the kinds of analyses completed for all run-types combined.

## next steps

Can I look at this for WR only? Include other LAD-based run designations?

Try the mixed effects modeling approach that Jereme recommended.

What happens if I limit the data to genetically-typed salmon but with (necessarily) reduced time series?

- Salvage data from 1993 through the end of the calendar year 2025; Afentoulis, V., G. Aasen, W. Griffiths, T. Celeste, T.D. Malinich, S. Siddiqui, and L. Damon. 2026. Fish Facilities Salvage Data, San Francisco, California, 1993-2025 ver 9. Environmental Data Initiative. <https://doi.org/10.6073/pasta/f1da5e94135634b7f601be6379dda133>

- Data from 3,004 acoustically-tagged Chinook tagged in 15 studies, over a 12-year period; Burdick, S.M., Raboin, M.J., Plumb, J.M., and Perry, R.W., 2026, Detection of Four Runs of Telemetry Tagged Juvenile Chinook Salmon in the Sacramento River and Georgiana Slough and Associated Flow Data from 2009 to 2022: U.S. Geological Survey data release, <https://doi.org/10.5066/P16VW2OM.>

Need to consider finer spatial scales: what do accumulation curves for individual sites look like?

What about including the whole year and limiting the time series to post-1992 (or thereabouts)?

Surely there is an applicable metric for the magnitude of a first flush event. Include that in one of these models!

What about other salmon data sources (eg trawls from Chipps Island)? What about other fish species?
