---
title: "salmon outmigrant phenology"
author: "Pete Nelson"
editor: visual
---

# setup

One package *not* available on CRAN: If necessary, get it using `remotes::install_github("doi-usgs/EflowStats@v5.2.0")`

```{r}
#| label: packages
#| message: false

# library(conflicted) # causes WAAAY more trouble than it solves!
library(tidyverse)
library(here)
library(readxl)
library(janitor) 
library(ggplot2)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
library(EflowStats)
library(fishualize)
library(tidybayes)
```

# load data

```{r}
#| label: seine-data
#| echo: true

# beach seine data
raw <- 
  read_rds(here("salmon/data/raw/raw.rds"))

# naming conventions
taxonomy <- 
  read_rds(
  here("salmon/data/raw/dt4.rds"))

# site information
sample_sites <- 
  read_rds(
    here("salmon/data/raw/dt5.rds"))

# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

```

# wrangle data

## choose sites

```{r}
#| label: limit-sites

# list of beach seine station codes identified minimally adequate for 
# temporal coverage
good <- 
  tibble(station = read_rds(here("salmon/data/processed/selected_site_codes.rds")))

# select data from those minimum effort sites and for complete water years (1976-10-01 onwards)
dat <-
  raw %>% 
  filter(station %in% good$station,
         # limit to data collected during complete water years
         date >= "1976-10-01") %>% 
  mutate(year = year(date),
         month = month(date)) %>% 
  relocate(region, .after = id_code) %>% 
  relocate(year:month, .before = date)

reg_labs <-
  tibble(region = factor(1:6),
         region_label = c("Lower Sac R", "N Delta", "Cen Delta", "S Delta",
                          "Lower SJR", "SF & SP Bays"))

# create monthly coverage indicator and retain region_code
coverage <- dat %>%
  distinct(station, year, month, region) %>%
  # 'tag' months sampled
  mutate(sampled = 1)

# create tibble of station_code and region_code for later use
sta_reg <-
  coverage %>%
  distinct(station, region) %>% 
  left_join(reg_labs, by = "region")

```

## water year types

Add data on *water year types* to data frame.

```{r}
#| label: wy-wyd

# add variables: water year and day of water year
# select data begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

dat <-
  dat %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = dataRetrieval::calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = EflowStats::get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  dplyr::select(c(station:water_temp, organism_code, fork_length, race_by_length, count)) %>% 
  # add 'sample_code' to identify each unique seine event
  group_by(date, time, station) %>% 
  mutate(sample_id = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(sample_id)

```

# analysis

## daily salmon counts

Calculate the accumulated number of salmon over the course of each water year and aggregate over all beach seine sites. Determine a cumulative percentage for each day over the entire year and for the 3-month period January, February, and March. Plot annual accumulation curves to show the pattern of juvenile outmigration for each water year, as well as for each Jan-Mar period for each year.

```{r}
#| label: counts

# daily counts
# pivot to simplify summing Chinook counts (including 0's) for each sample unit
seine <-
  dat %>% 
  pivot_wider(
    # relies on 'sample_id' to distinguish among sample units, but retains environmental data
    id_cols = c(sample_id:time, water_temp),
    # columns for all fish spp (only CHN needed for this analysis)
    names_from = organism_code,
    # log number of each speciesxfork_length for each seine (sample_id)
    values_from = count,
    # if no matches, enter '0'
    values_fill = 0,
    # sum counts if multiple matches
    values_fn = sum
  ) %>% 
  dplyr::select(c(sample_id:water_temp, CHN))

### daily counts -----
# aggregate to daily totals bc freq >>1 seine per day
# note that this will obscure potential differences among locations

daily <-
  seine %>%
  group_by(wy, month, wd) %>%
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent------
# calc cumulative percentage for each wy
daily_cum <-
  daily %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ungroup()

# calc cumulative percentage for each wy but limited to the period from Jan 1-Mar 31
# Jan 1 is wd 93; Mar 31 is wd 182 (leap years, it's 183 but I'm going to ignore this)
daily_cum_limited <-
  daily %>% 
  filter(month <= 3) %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ungroup()

```

```{r}
#| label: accum-curves
#| #| message: false

# initial tests...no output desired because used all months and these data are poor for many years/sites
# changed the colors where the older years are darker and newer years are lighter to see the phenological shifts in timing 

daily_cum %>% 
  arrange(wy, wd) %>% 
ggplot(aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  # add 1977 on top in red
  geom_line(data = subset(daily_cum, wy == 1977), 
            color = "red", 
            linewidth = 0.7) +
  xlim(0, 300) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  annotate("text", x = 250, y = 30,
           label = "1977", color = "red", fontface = "bold") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "data from January - December, 1977-2024",
       color = "water year") +
  theme_bw()

ggsave(here("salmon/plots/accum_allmonths.png"))

# "outlier" is 1977

# plotted data from 1992 (when seines were conducted monthly year round) to 2024; not used

ggplot(daily_cum %>% 
         filter(wy >= 1992),
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  xlim(0, 300) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "data from January - December, 1992-2024",
       color = "water year") +
  theme_bw()

# THIS is the approach that we prob want to use
# limit the data to January - March: data = daily_cum_limited
ggplot(daily_cum_limited,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  # add 1977 on top in red
  geom_line(data = subset(daily_cum_limited, wy == 1977), 
            color = "red", 
            linewidth = 0.7) +
  xlim(0, 300) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  annotate("text", x = 190, y = 30,
           label = "1977", color = "red", fontface = "bold") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "data from January - March only, 1977-2024",
       color = "water year") +
  theme_bw()

ggsave(here("salmon/plots/accum_jan2mar.png"))
```

## benchmarks

Determine the day of each water year when 10%, 50% and 90% of the total annual numbers are reached.

```{r}
#| label: benchmarks

## ID benchmarks ----
# day of water year when the 10%, 50% and 90% of total annual numbers are reached

# using 10%, 50%, and 90% accumulation benchmarks (arbitrary)
benchmarks <- c(10, 50, 90)

# gen df of water year days when threshold reached for each year
percentile_days <- 
  daily_cum %>%
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cum_percent >= 10]),
    day_50 = min(wd[cum_percent >= 50]),
    day_90 = min(wd[cum_percent >= 90]),
    .groups = "drop"
  )

# gen df of water year days when threshold reach for each year, using data from
# January - March only
percentile_days_limited <- 
  daily_cum_limited %>%
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cum_percent >= 10]),
    day_50 = min(wd[cum_percent >= 50]),
    day_90 = min(wd[cum_percent >= 90]),
    .groups = "drop"
  )

# pivot data to long format to facilitate graphics
pctl_days <-
  percentile_days %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

pctl_days_limited <-
  percentile_days_limited %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

```

## water year type

Add the water year type data to the percentile data for analysis and graphics.

```{r}
#| label: wy-type

## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

pctl_days_wyt <-
  wy_type %>% 
  # select Sac watershed, excludes SJR watershed
  filter(watershed == "Sacramento Valley") %>% 
  # merge w simple df, includes 1977 data 
  right_join(pctl_days,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, "W", "AN", "BN", "D", "C"),
         # center wy to facilitate model convergence
         wy_centered = wy - 1977) %>% 
  relocate(wy_centered, .after = wy)

# make 'wyt' an ordered factor
pctl_days_wyt <- pctl_days_wyt %>%
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      # must be low to high, where C<D<BN<AN<W
                      ordered = TRUE))

write_rds(pctl_days_wyt, 
          here("salmon/data/processed/pctl_days_wyt.rds"))

```

## migration timing

Plot the water year days when each benchmark (10, 50, or 90%) is reach for each water year.

```{r}
#| label: wy-effect

# figure shows the day of the water year for each benchmark (10, 50, 90%) for each year

benchmark_trends_limited <-
  pctl_days_limited %>% 
  ggplot(
    # color points according to threshold (10, 50, 90%)
    aes(x = wy, y = wd, color = percentile)
  ) +
  geom_point(size = 4) +
  scale_color_fish(discrete = TRUE,
                  option = "Oncorhynchus_tshawytscha",
                  alpha = 0.6) +
  geom_smooth(data = pctl_days_limited[4:147,],
              alpha = 0.2,
              # show linear trends
              method = "lm",
              se = TRUE) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "data from January-March only; 1977 excluded from trend analyses",
       x = "Water Year",
       y = "Day of the Water Year",
       color = "Percentile") +
  theme_bw()

benchmark_trends_limited

```

## models

### year effect

Testing for a water *year* effect, focusing first on the *day10* benchmark, when 10% of each annual total has been reached.

```{r}
#| label: m3-day10

percentile_days_limited

# first including data from 1977
# reaching the 10% benchmark
m3.day10.wy <- 
  brm(data = percentile_days_limited, # use the percentile_days_limited data
      # use a gaussian (normal) distribution
      family= gaussian,
      # specify the model
      day_10 ~ 1 + wy,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      # set seed so team gets the same results when running the model
      seed = 42,
      file = here("salmon/results/m3.day10.wy"),
      file_refit = "on_change")

# run posterior predictive check
pp_check(m3.day10.wy)

write_rds(m3.day10.wy, here("salmon/results/m3.day10.wy.rds"))

m3.day10.wy_robust <- 
  brm(data = percentile_days_limited,
      family = student,  # <--- change from gaussian to student
      day_10 ~ 1 + wy,
      # apply loo_moment_match on problematic observations bc crazy 1977 outlier
      save_pars = save_pars(all = TRUE),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/m3.day10.wy_robust"),
      file_refit = "on_change")
# looks a bit better!
pp_check(m3.day10.wy_robust)

# calculate LOO for the original Jan-Mar Gaussian model
m3.day10.wy <- add_criterion(m3.day10.wy, criterion = "loo")
# warning about "Pareto values > 0.7" happens when one data point (like 1977) is so influential that the model changes drastically when it's removed

# calculate LOO for the new "robust" (Student-t) model
m3.day10.wy_robust <- add_criterion(m3.day10.wy_robust, criterion = "loo")

# compare the two day10 models
loo_compare(m3.day10.wy, m3.day10.wy_robust, criterion = "loo")

# notes:
# the performance difference between the models is very modest: 
# the rule of thumb is: consider one model "clearly better" if the elpd_diff is more than twice the se_diff, but my difference is 2.8, and my error is 1.5, so both models are performing similarly

m3.day10.wy_robust$criteria$loo
m3.day10.wy$criteria$loo

# the robust model (using Student-t distribution) seems to perform slightly better, 
# but the improvement could plausibly be attributed to noise

# let's plot the k values for each model
plot(m3.day10.wy$criteria$loo)
plot(m3.day10.wy_robust$criteria$loo)
# PSIS diagnostic plots confirm that both models are pretty good, but also 
# show how nicely the 'robust' model handles the extreme datum from 1977

summary(m3.day10.wy_robust)
plot(m3.day10.wy_robust)

m3.day10.wy_robust %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
```

The "robust" model for the *day10* benchmark provides strong support for the hypothesis that migration, particularly the earlier phase of the process, has shifted earlier in the water year by approximately 0.35 days each year from 1997-2024.

The code below models comparable relationships to the previous code, but uses the *day50* and *day90* benchmarks. This code needs to be reviewed for accuracy though.

```{r}
#| label: m3-day50+

# check this code!

# reaching the 50% benchmark
m3.day50.wy <- 
  brm(data = percentile_days_limited, 
      # Choose a gaussian (normal) distribution
      family= gaussian,
      day_50 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/m3.day50.wy"),
      file_refit = "on_change")

# reaching the 90% benchmark
m3.day90.wy <- 
  brm(data = percentile_days_limited, 
      family= gaussian,
      day_90 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/m3.day90.wy"),
      file_refit = "on_change")

summary(m3.day50.wy)
summary(m3.day90.wy)

plot(m3.day50.wy)
plot(m3.day90.wy)

# calc the prob that the slope is less than 0

m3.day10.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m3.day50.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m3.day90.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

```

I tested multiple models to explore the timing of juvenile salmon outmigration, represented by the day of the water year when three benchmarks had been reached over the course of the time series. Each model used a Gaussian distribution with *water year* as the predictor and the day of the water year when those benchmarks–10, 50, and 90--are reached as the response variable: *day10, day50, day90*. I limited the data to those gathered during January, February and March because these were sampled consistently (albeit imperfectly) in every year. Early outmigrants entering the Delta in December or earlier would be missed, but presumably the most important early numbers would be reasonably well represented.

These results present strong evidence for a *year* effect (i.e., model slopes were different from zero, however, the posterior distribution for the *day90* model overlaps 0 slightly (95% CIs: -0.73, 0.05).

-For the 10% accumulation model (*day10*), there was an estimated slope of -0.69 days per year and the probability that the slope is less than 0 equals 0.999.

-For the 50% accumulation model (*day50*), there was an estimated slope of -0.70 days per year and the probability that the slope is less than 0 equals 1.000.

-For the 90% accumulation model (*day90*), there was an estimated slope of -0.33 days per year and the probability that the slope is less than 0 equals 0.955.

### water year type effect

I also attempted to determine if there was an effect of water year *type* on the timing of Chinook outmigration. I tested several models that included both *water year* and *wyt* (water year type) as explanatory variables.

```{r}
#| label: wyt-effect
## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

pctl_days_wyt <-
  wy_type %>% 
  # select Sac watershed, excludes SJR watershed
  filter(watershed == "Sacramento Valley") %>% 
  # merge w simple df, includes 1977 data 
  right_join(pctl_days,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, "W", "AN", "BN", "D", "C"),
         # center wy to facilitate model convergence
         wy_centered = wy - 1977) %>% 
  relocate(wy_centered, .after = wy)

# make 'wyt' an ordered factor
pctl_days_wyt <- pctl_days_wyt %>%
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      # must be low to high, where C<D<BN<AN<W
                      ordered = TRUE))

write_rds(pctl_days_wyt, 
          here("salmon/data/processed/pctl_days_wyt.rds"))

### wy boxplot -----
pctl_days_wyt %>% 
  filter(percentile == "day_10",
         # no data yet re wyt for 2025
         wy <= 2024) %>% 
  ggplot(aes(x = wyt, y = wd, group = wyt)) +
  geom_boxplot() +
  labs(title = "Timing of 10% Annual Accumulation Benchmark",
       x = "Water Year Type",
       y = "Day of the Water Year") +
  theme_bw()

### m1 ----
# run wy + wyt model with the centered variable, where wyt is ordered & monotonic
# 2025 not included
m1.day10.wy.wyt <- 
  brm(data = pctl_days_wyt %>% 
        filter(wy <= 2024),
      family = gaussian,
      wd ~ 0 + wy_centered + mo(wyt), 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, 
      seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("salmon/results/m1.day10.wy.wyt"),
      file_refit = "on_change"
  )

summary(m1.day10.wy.wyt)
plot(m1.day10.wy.wyt)
# Rhat is perfect
# water year (centered) effect looks solid 
# posterior distributions look great (uniform)
# mcmc look great too
# estimates for year type are wonky--reflects the data! 
# strong effect of 'year type' as a monotonic, ordered factor. cool!
# is there an interaction effect?
# should also include some semi-informed priors?

# compute estimated effect for each wyt level
# get predicted effect of each wyt level, holding wy_centered at 0
ce <- conditional_effects(m1.day10.wy.wyt, effects = "wyt", re_formula = NA)

# extract the estimate for each level, plus 95% CIs
wyt_effects <- ce$wyt[, c("wyt", "estimate__", "lower__", "upper__")]

# round for readability
wyt_effects$estimate <- round(wyt_effects$estimate__, 2)

wyt_effects %>% 
  dplyr::select(-estimate__) %>% 
  mutate("l-95% CI" = lower__,
         "u-95% CI" = upper__,
         .keep = "unused")

# note that these are scaled effects and can't be translated directly into days of the water year! here's an attempt to translate monotonic effects into predicted timing
# Define levels of water year type
levels_wyt <- c("C", "D", "BN", "AN", "W")

# Predicted wd at reference wy_centered = 0
preds <- fitted(m1.day10.wy.wyt, newdata = data.frame(
  wyt = levels_wyt,
  wy_centered = 0
))

# Assemble into a table
predicted_wd <- data.frame(
  wyt = levels_wyt,
  estimate = round(preds[, "Estimate"], 1),
  l_95 = round(preds[, "Q2.5"], 1),
  u_95 = round(preds[, "Q97.5"], 1)
)

predicted_wd

### m2 ----
# run wy * wyt model with the centered variable, 2025 not included
m2.day10.wy.wyt <- 
  brm(data = pctl_days_wyt %>% 
        filter(wy <= 2024),
      family = gaussian,
      wd ~ 0 + wy_centered * wyt, 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, 
      seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("salmon/results/m2.day10.wy.wyt"),
      file_refit = "on_change"
  )

summary(m2.day10.wy.wyt)
plot(m2.day10.wy.wyt)

### m3 ----
# run wy * wyt model with the centered variable, 2025 not included, replace Gaussian w Students-t distribution
m3.day10.wy.wyt <- 
  brm(data = pctl_days_wyt %>% 
        filter(wy <= 2024),
      family = student,
      wd ~ 0 + wy_centered * wyt, 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, 
      seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("salmon/results/m3.day10.wy.wyt"),
      file_refit = "on_change"
  )

# to be consistent, I should prob designate this model as 'robust', but can't be bothered
summary(m3.day10.wy.wyt)
plot(m3.day10.wy.wyt)

# posterior predictive check (see density overlay figures)
pp_check(m1.day10.wy.wyt)
pp_check(m2.day10.wy.wyt)
pp_check(m3.day10.wy.wyt)

# calculate LOO for the original water_year + water_year_type Gaussian model
m1.day10.wy.wyt <- add_criterion(m1.day10.wy.wyt, criterion = "loo")

# calculate LOO for the water_year * water_year_type Gaussian model
m2.day10.wy.wyt <- add_criterion(m2.day10.wy.wyt, criterion = "loo")

# calculate LOO for the water_year * water_year_type Student-t "robust" model
m3.day10.wy.wyt <- add_criterion(m3.day10.wy.wyt, criterion = "loo")

# compare the three day10 wyt models with interaction terms
loo_compare(m1.day10.wy.wyt, m2.day10.wy.wyt, m3.day10.wy.wyt, criterion = "loo")

# no real difference in model performance among the 3...

m1.day10.wy.wyt$criteria$loo
m2.day10.wy.wyt$criteria$loo
m3.day10.wy.wyt$criteria$loo

# let's plot the k values for each model (cause I like these figures)
plot(m1.day10.wy.wyt$criteria$loo)
plot(m2.day10.wy.wyt$criteria$loo)
plot(m3.day10.wy.wyt$criteria$loo)
# everything well below 0.7...

```
