---
title: "salmon outmigrant phenology"
author: "Pete Nelson"
format: 
  html:
    embed-resources: true
execute:
  execute-dir: project
editor: visual
---

# introduction

The intent here is to explore the beach seine data from the DJFMP dataset for patterns in the presence and abundance of juvenile Chinook salmon in the lower Sacramento River and the Delta. These data were first collected in 1976 and the most recent data are from water year 2024.

# setup

One package *not* available on CRAN: If necessary, get it using `remotes::install_github("doi-usgs/EflowStats@v5.2.0")`

```{r}
#| label: packages
#| message: false

# library(conflicted) # causes WAAAY more trouble than it solves!

library(here)
library(readxl)
library(janitor) 
library(ggplot2)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
library(EflowStats)
library(fishualize)
library(tidybayes)
library(tidyverse)

```

# load data

```{r}
#| label: seine-data
#| echo: true

# beach seine data ('raw' includes all spp)
raw <- 
  read_rds(here("salmon/data/raw/raw.rds"))

# naming conventions
taxonomy <- 
  read_rds(
  here("salmon/data/raw/dt4.rds"))

# site information
sample_sites <- 
  read_rds(
    here("salmon/data/raw/dt5.rds"))

# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

```

## storm data

Let's pull in Rosie's storm data—I think this includes *first flush* events, both timing and magnitude.

```{r}
#| label: storm-data

load(here("data/processed/storms/StormData.RData"))

# use 'Firststorms'
storms <-
  Firststorms %>% 
  arrange(WY) %>% 
  mutate(wyd = get_waterYearDay(Date))

write_rds(storms, here("salmon/data/processed/storms.rds"))

```

# wrangle data

## choose sites

There are beach seine data from a large number of sites, but the consistency with which these sites were samped varies widely. In a separate document, I provide the details used to select sites that were sampled consistently from water year 1976-2024. Even at these sites, however, there are missing data, particularly for the months April-December prior to 1992.

```{r}
#| label: limit-sites

# list of beach seine stations identified minimally adequate for 
# temporal coverage (tibble includes region and region_label)
sites <- 
  read_rds(here("salmon/data/processed/sites.rds"))

# select data from those minimum effort sites and for complete water years (1976-10-01 onwards)
dat <-
  raw %>% 
  filter(station %in% sites$station,
         # limit to data collected during complete water years
         date >= "1976-10-01") %>% 
  mutate(year = year(date),
         month = month(date)) %>% 
  relocate(region, .after = id_code) %>% 
  relocate(year:month, .before = date)

# create monthly coverage indicator and retain region_code
coverage <- dat %>%
  distinct(station, year, month, region) %>%
  # 'tag' months sampled
  mutate(sampled = 1)

```

## water year types

In California, *water year types* are classifications (Wet, Above Normal, Below Normal, Dry, and Critical) that are used to categorize hydrological conditions over a 12-month period (October 1–September 30), based, for the Sacramento and San Joaquin valleys, on unimpaired runoff. These designations are used for making water management decisions and probably are not the best environmental measures available for understanding salmon ecology, but they are used by managers and offer a reasonable starting point for exploring the relationship between environment and salmon migration. Additional environmental metrics will be explored in the future.

Add data on *water year types* to the data:

```{r}
#| label: wy-wyd

# add variables: water year and day of water year
# select data begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

dat <-
  dat %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = dataRetrieval::calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = EflowStats::get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  dplyr::select(c(station:water_temp, organism_code, fork_length, race_by_length, count)) %>% 
  # add 'sample_code' to identify each unique seine event
  group_by(date, time, station) %>% 
  mutate(sample_id = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(sample_id)

```

# analysis

## accumulation curves

Calculate the accumulated number of salmon over the course of each water year and aggregate over selected beach seine sites. Determine a cumulative percentage for each day over the entire year and for the 3-month period January, February, and March. Plot annual accumulation curves to show the pattern of juvenile outmigration for each water year, as well as for each January-March period for each year. The truncated data are used as these months were sampled consistently throughout the time series; sampling effort during the remaining months, April-December was limited prior to 1992.

```{r}
#| label: counts

# daily counts
# pivot to simplify summing Chinook counts (including 0's) for each sample unit
seine <-
  dat %>% 
  pivot_wider(
    # relies on 'sample_id' to distinguish among sample units, but retains environmental data
    id_cols = c(sample_id:time, water_temp),
    # columns for all fish spp (only CHN needed for this analysis)
    names_from = organism_code,
    # log number of each speciesxfork_length for each seine (sample_id)
    values_from = count,
    # if no matches, enter '0'
    values_fill = 0,
    # sum counts if multiple matches
    values_fn = sum
  ) %>% 
  dplyr::select(c(sample_id:water_temp, CHN)) %>% 
  left_join(sites, by = "station") %>% 
  dplyr::select(-region) %>% 
  relocate(region_label, .after = location)

### daily counts -----
# aggregate to daily totals bc freq >>1 seine per day

# aggregated across all selected sites
daily_delta <-
  seine %>%
  group_by(wy, month, wd) %>%
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  )

# aggregated by region
daily_regional <-
  seine %>% 
  group_by(region_label, wy, month, wd) %>% 
  summarise(
    daily_total = sum(CHN, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent------
# limited to the period from Jan 1-Mar 31
# calc cumulative percentage for each wy during that 3-month period, delta-wide
delta_cumul <-
  daily_delta %>% 
  filter(month <= 3) %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/delta_cumul.rds"))

# cumulative totals and percentages for Jan-Mar by region
regional_cumul <-
  daily_regional %>% 
  filter(month <= 3) %>% 
  group_by(wy, region_label) %>% 
  arrange(region_label, wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(region_label, wy, month, wd) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/processed/regional_cumul.rds"))

```

Plot the accumulation curves.

```{r}
#| label: annual-accumulation
#| message: false
#| warning: false

# annual (whole year) accumulation curves, lumping data from all sites
daily_delta %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  ggplot(aes(x = wd,
           y = cumul_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  xlim(0, 300) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  # highlight 1977 data
  annotate("text", x = 250, y = 30,
           label = "1977", color = "red", fontface = "bold") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Delta-Wide Accumulation Curves",
       subtitle = "data from 1977-2024",
       color = "water year") +
  theme_bw()

```

```{r}
#| label: 3month-accumulation

regional_cumul %>% 
  arrange(region_label, wy, wd) %>%
  mutate(region_label = fct_relevel(region_label, 
                                     "Lower Sacramento River", "North Delta", 
                                     "Central Delta", "South Delta")) %>% 
  ggplot(aes(x = wd,
             y = cumul_percent,
             color = wy,
             group = wy)) +
  geom_line() +
  xlim(50, 200) +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako") +
  labs(x = "Day of Water Year",
       y = "Cumulative % of 3-Month Total", 
       title = "Juvenile Chinook Regional Accumulation Curves",
       subtitle = "January - March, 1977-2024",
       color = "Water Year") +
  facet_wrap(~region_label, nrow = 2) +
  theme_bw()

```

## benchmarks

Determine the day of each water year when 10%, 50% and 90% of the total annual numbers are reached.

```{r}
#| label: benchmarks
#| message: false

## ID benchmarks ----
# day of water year when the 10%, 50% and 90% of total annual numbers--accumulation benchmarks

# for entire year, entire delta
annual_benchmarks <-
  daily_delta %>% 
  group_by(wy) %>% 
  arrange(wy, wd) %>% 
  mutate(
    annual_total = sum(daily_total),
    cumul_total = cumsum(daily_total),
    cumul_percent = 100 * cumul_total / annual_total
  ) %>% 
  arrange(wy, month, wd) %>% 
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

three_month_delta_benchmarks <-
  delta_cumul %>% 
  arrange(wy, wd) %>% 
  group_by(wy) %>% 
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

three_month_regional_benchmarks <-
  regional_cumul %>% 
  arrange(region_label, wy, wd) %>% 
  group_by(region_label, wy) %>% 
  summarise(
    day_10 = min(wd[cumul_percent >= 10]),
    day_50 = min(wd[cumul_percent >= 50]),
    day_90 = min(wd[cumul_percent >= 90]),
    .groups = "drop"
  ) %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

```

## water year type

Add the water year type data (loaded previously) to the percentile data for analysis and graphics.

```{r}
#| label: wy-type

## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

three_month_regional_benchmarks <-
  wy_type %>% 
  # select Sac watershed; exclude SJR watershed
  filter(watershed == "Sacramento Valley") %>% 
  # merge with benchmark data
  right_join(three_month_regional_benchmarks,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, 
                           c("C", "D", "BN", "AN", "W")),
         # center wy to facilitate model convergence
         wy_meanCentered = wy - mean(wy)) %>% 
  relocate(wy_meanCentered, .after = wy)

# make 'wyt' an ordered factor
three_month_regional_benchmarks <- 
  three_month_regional_benchmarks %>%
  mutate(wyt = factor(wyt, levels = c("C", "D", "BN", "AN", "W"), 
                      # must be low to high, where C<D<BN<AN<W
                      ordered = TRUE)) %>% 
  write_rds(
    here("salmon/data/processed/three_month_regional_benchmarks.rds"))


```

## migration timing

Plot the water year days when each benchmark (10, 50, or 90%) is reach for each water year.

```{r}
#| label: wy-effect

# quick data qaqc check (shouldn't be any NAs)
unique(three_month_regional_benchmarks$region_label)

# 10% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_10 benchmark
  filter(percentile == "day_10") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "10% of 3-Month Total Reached (day of water year")

ggsave(here("salmon/plots/benchmark_10pct.png"))

# 50% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_50 benchmark
  filter(percentile == "day_50") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "50% of 3-Month Total Reached (day of water year")

ggsave(here("salmon/plots/benchmark_50pct.png"))

# 90% benchmark
three_month_regional_benchmarks %>%
  # limit to the day_90 benchmark
  filter(percentile == "day_90") %>%
  mutate(
    # relevel region_label to establish factor order
    region_label = factor(
      region_label,
      levels = c("Lower Sacramento River",
                 "North Delta",
                 "Central Delta",
                 "South Delta")
    )
  ) %>%
  ggplot(
    aes(x = wy, y = wd, 
        # color by region
        color = region_label)
  ) +
  geom_point(size = 4) +
  # fish colors!
  scale_color_fish(discrete = TRUE,
                   option = "Oncorhynchus_tshawytscha",
                   alpha = 0.6) +
  # add trend lines
  geom_smooth(alpha = 0.2,
              method = "lm",
              se = TRUE) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ region_label, ncol = 2) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "data from January - March only",
       x = "Water Year",
       y = "90% of 3-Month Total Reached (day of water year")

ggsave(here("salmon/plots/benchmark_90pct.png"))


```

## models

### change over time and water year type effects

Testing for potential effects of water year (i.e., change over time), water year type, regional differences, and differences in the three benchmarks (is there a difference in the slopes).

```{r}
#| label: bayes1
#| results: false
#| message: false
#| warning: false

# ensure that we've got a nested directory
output_dir <- here("salmon", "results")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# I've used a Gaussian distribution (at least for now) because the distributions for each
# benchmark look "normalish." See
three_month_regional_benchmarks %>%
  ggplot(aes(wd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")
# I'll do a pp_check of the model; if this shows the model giving wd=60 or wd=250 with
# non-trivial probabilities, I'll have to swtich to Beta or truncated normal distribution

bmod1 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = gaussian,
      wd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      # set seed
      seed = 42,
      # brms automatically appends '.rds'
      file = here("salmon/results/bmod1"),
      file_refit = "on_change"
      )

pp_check(bmod1) # looks good!
(bmod1_sum <- summary(bmod1)) # see analysis
plot(bmod1) # looks good
conditional_effects(bmod1, "wy_meanCentered:percentile") # doesn't add much

```

All `Rhat = 1.00`, so the model converged cleanly. We have effective sample sizes in the low thousands and `n = 507`, so we can take the estimates seriously.

- `Intercept` (125.59, CI \[122.13, 128.89\]): Over the entire time series, in the Central Delta, the day_10 benchmark is reached roughly on wyd 126 or about February 2.

- `wy_meanCentered` (-0.44, CI \[-0.65, -0.24\]): The day_10 benchmark is shifting earlier by about 0.44 days per year, and, over the roughly 50-year span of our time series, that's about a 20 day shift earlier in the year. The credible interval excludes 0, so this is a real and significant trend!

- The percentile main effects (`percentileday_50` +21.76 and `percentileday_90` +44.74) are no surprise–the way that these are determined ensures that these follow the 10% benchmark day, but the magnitudes tell us how spread out the distribution is in a typical year.

- Regional effects (baseline = Central Delta): The Lower Sacramento River is about 8 days earlier than the Central Delta, but benchmarks reached in the North and South Deltas are indistinguishable from that in the Central Delta (both CIs include zero). So, not surprisingly, fish reach the Lower Sacramento River sites first, and then the three Delta regions. (Still, it's comforting to see things work "the way they should.")

- Percentile x year interactions:

  - `wy_meanCentered:percentileday_50` = -0.09, CI \[0.12, -0.32\]

  - `wy_meanCentered:percentileday_90` = 0.13, CI \[0.12, -0.11\]

- Both CIs cross zero so we can't say that these slopes are *different from* the day_10 slope; rather, the day_10 slope is credibly negative (-0.44, definitively earlier) and other slopes are indistinguishable from the first.

- `mowyt` (-0.96, CI \[-1.77, -0.17\]): As you move from the driest (C) to wettest (W) water year types, `wd` decreases by about 1 day total. So, wet years reach the benchmarks very slightly earlier than dry years, holding everything else constant. This is a very (very) modest effect compared to the change over time (years) and the difference between river sites and Delta sites.

- `mowyt:wy_meanCentered` (+0.09, CI \[0.03, 0.03\]): Year type effects interact with year such that the gap between wet and dry years has been *narrowing* over time. While the effect size is small, it's credibly non-zero.

- Monotonic simplex parameters: The monotonic effect among year types is distributed very evenly, so no single step (e.g., BN to AN) dominates, and the assumption of monotonicity is reasonable. Of course, the way these are determined, it *ought* to be!

- `sigma` (14.62): Individual observations of `wd` vary by about 15 days around the model's prediction, after accounting for year, percentile (benchmark), region and water year type. Consider this the unexplained year-to-year and within-region variance.

**Summary** The entire distribution of the wd benchmarks (10, 50th, 90th percentiles) is moving more than 0.4 days earlier each year, and the shift is statistically credible. The three benchmarks are shifting in parallel rather than diverging or converging. Fish arrive in the Lower Sacramento River about a week prior to their arrival in the Delta, and timing to reach sites within the Delta are indistinguishable from each other. Arrival times are slighly earlier during wet years than in dry years, but that gap has narrowed over time.

I ran the same model using a Students t-distribution; based mostly on the prior posterior check, I don't see that there's a significant improvement over the original, `bmod1`.

```{r}
#| label: robust
bmod2 <- 
  brm(data = three_month_regional_benchmarks %>% 
        rename(region = region_label),
      family = student,
      wd ~ 1 + wy_meanCentered * percentile + wy_meanCentered * mo(wyt) + region,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod2"),
      file_refit = "on_change"
      )

pp_check(bmod2) # looks good!
(bmod2_sum <- summary(bmod2)) # see analysis
plot(bmod2) # looks good
conditional_effects(bmod2, "wy_meanCentered:percentile") # doesn't add much
```

### linear approach

```{r}
#| label: linear-models


library(dplyr)
library(tidyr)
library(purrr)
library(broom)

models <- three_month_regional_benchmarks %>%
  group_by(percentile) %>%
  nest() %>%
  mutate(
    fit     = map(data, ~ lm(wd ~ wy_meanCentered + wyt + region_label, data = .x)),
    coefs   = map(fit, tidy, conf.int = TRUE),
    summary = map(fit, glance)
  )

# coefficient table across percentiles
# note: wyt.L is the linear trend;
# does wd go up or down with wyt from drier to wetter
models %>% 
  dplyr::select(percentile, coefs) %>% 
  unnest(coefs)

# model-level fit stats (R^2, AIC, etc.)
models %>% 
  dplyr::select(percentile, summary) %>% 
  unnest(summary)

```

#### notes

- `wyt.L` is the linear trend across the ordered levels — "as you move from drier to wetter years, does `wd` go steadily up or steadily down?" A positive coefficient means `wd` rises with wetness; negative means it falls. This is usually the one you care about most, and if it's the only significant term, the effect is essentially monotonic in wetness.

- `wyt.Q` captures curvature — a U-shape or inverted-U. Positive means the middle categories (BN, AN) have *lower* `wd` than the extremes (C and W); negative means the middle is *higher* than the extremes. Significant `.Q` with non-significant `.L` would mean "extremes behave alike, middle is different."

- `wyt.C` is an S-shape — `wd` rises then falls then rises (or vice versa) as you move through the ordering. Rarely interpretable on its own; significance here usually means the spacing between levels isn't smooth.

- `wyt^4` is an even wigglier shape with three bends. Almost never substantively meaningful — if it's significant, it's usually a sign the ordered-factor assumption isn't quite right.

Doesn't really add anything of value to the Bayesian models above.

### first flush effect

```{r}
#| label: ff-data


storms <- 
  read_rds(here("salmon/data/processed/storms.rds")) %>% 
  mutate(ff_wyd = wyd,
         .keep = "unused")
# no first flush storms for 1977, 2020, 2022 and 2025--filter(is.na(WY))

ff_data <- 
  three_month_regional_benchmarks %>% 
  left_join(storms,
            join_by(wy == WY),
            keep = FALSE) %>% 
  dplyr::select(wy, wy_meanCentered, ff_wyd, percentile, wd, SAC, region = region_label) %>% 
  filter(!is.na(ff_wyd))


```

Let's see if there are any patterns.

```{r}
#| label: plots


ff_data %>%
  ggplot(aes(wd)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ percentile, scales = "free")

ff_data %>% 
  ggplot(aes(x = wy, y = ff_wyd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  theme_bw()

# this is kinda crazy! water ops?
summary(lm(data = ff_data, ff_wyd ~ wy))

ff_data %>% 
  ggplot(aes(x = wy, y = wd, color = percentile)) +
  geom_point() +
  theme_bw()

```

```{r}
#| label: ff-models


bmod3 <- 
  brm(data = ff_data,
      # 1st interaction lets slope of wd differ across benchmarks
      # I expect that day_10 will track ff closely but day_90 not so much
      wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered,
      family = gaussian,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod3"),
      file_refit = "on_change")

pp_check(bmod3) # looks good!
(bmod3_sum <- summary(bmod3)) # see analysis
plot(bmod3) # looks good
conditional_effects(bmod3, "ff_wyd:percentile") # nice showing how ff matters more early

bmod4 <- 
  brm(data = ff_data,
      # within yr obs may not be independent?
      wd ~ 1 + ff_wyd * percentile + region + wy_meanCentered + (1 | wy),
      family = gaussian,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/bmod4"),
      file_refit = "on_change")

pp_check(bmod4) # looks good!
(bmod4_sum <- summary(bmod4)) # see analysis
plot(bmod4) # looks good
conditional_effects(bmod4, "ff_wyd:percentile")

```

SCRATCH below this
