---
title: "FirstFlush_Salmon"
author: "Pete Nelson"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

## packages

One package **not** available on CRAN: If necessary, `remotes::install_github("doi-usgs/EflowStats@v5.2.0")`.

```{r load_packages, include=FALSE}

library(here)
library(tidyverse)
library(readxl)
library(janitor) 
library(ggplot2)
library(MASS)
library(brms)
library(chron)
library(gridExtra)
library(dataRetrieval)
library(EflowStats)
library(fishualize)
library(tidybayes)
```

## beach seine data

Insert -\> Citation -\> Zotero

[@sturrock2019]

Beach seine data from the Delta Juvenile Fish Monitoring Program (DJFMP). *United States Fish and Wildlife Service, J. Speegle, C. Macfarlane, E. Huber, D. Goodman, K. Erly, and R. Cook. 2026. Interagency Ecological Program: Over four decades of juvenile fish monitoring data from the San Francisco Estuary, collected by the Delta Juvenile Fish Monitoring Program, 1976-2025 ver 14. Environmental Data Initiative.* https://doi.org/10.6073/pasta/ad35c80cafd87ba928d3f37fa124ccaf (Accessed 2026-03-11).

Note: Look at this data set too eventually. *United States Fish and Wildlife Service, J. Kirsch, M. Marshall, C. Johnston, A. Arrambide, and J. Speegle. 2025. Interagency Ecological Program and US Fish and Wildlife Service: Beach Seine Efficiency Study, collected by the Delta Juvenile Fish Monitoring Program, 2013-2018 ver 2. Environmental Data Initiative.* https://doi.org/10.6073/pasta/818b8bf1d93b81cc7c5124f9df81ce01 (Accessed 2026-03-11).

```{r load_data, eval=FALSE}
# beach seine data
raw <- 
  read_rds(here("salmon/data/raw/raw.rds"))

# naming conventions
taxonomy <- 
  read_csv(
  here("salmon", "data", "raw", "taxonomy.csv"))

# site information
sample_sites <- 
  read_csv(
    here("salmon", "data", "raw", "sample_sites.csv"))

# pull all IEP fish codes out to help with filtering 
# for fish data only (ie no inverts)
fish_codes <- 
    taxonomy %>% 
    filter(Phylum == "Chordata",
           # remove records where Species is NA
           # but retains eg "Bass unidentified"
           Species != "<NA>") %>%
    distinct(IEPFishCode)

```

## site selection

Refer to 'beach seine effort.qmd' for information on site selection.

```{r}

# list of beach seine station codes identified minimally adequate for 
# temporal coverage
good <- 
  tibble(station = read_rds(here("salmon/data/processed/selected_site_codes.rds")))

# select data from those minimum effort sites and for complete water years (1976-10-01 onwards)
dat <-
  raw %>% 
  filter(station %in% good$station,
         # limit to data collected during complete water years
         date >= "1976-10-01")
```

## data reduction+

Add water year and day of water year to the data frame.

Notes:

-   trawl data (esp Chipps Island)

-   try 1992+

-   check time window (does NOT seem to be have been effective)

-   spatial partitioning by rm, bin by region or subregion

-   beach seine data should still be helpful for teasing out the source of these juveniles (Am R, Feather, SJR, etc)

-   seine data useful for looking at the Delta rearing duration too

-   don't worry now about run type

```{r}
# add water year and day of water year
# select data begining with WY 1976 
# (Oct 1 1976 - Sep 30 1977) and ending with WY 2024 (Oct 1 2024 - end of 
# data set (July 31, close enough to the proper end of the WY for our purposes))

dat <-
  dat %>% 
  # add water year (wy) and day of water year (wd) variables
  mutate(
    # water year
    wy = dataRetrieval::calcWaterYear(date),
    # day of water year (wd), wyMonth=10L bc water year begins Oct 1
    wd = EflowStats::get_waterYearDay(date, wyMonth = 10L)) %>% 
  # reorg & select columns
  relocate(wy:wd, .after = date) %>% 
  select(c(station:water_temp, organism_code, fork_length, race_by_length, count)) %>% 
  # add 'sample_code' to identify each unique seine event
  group_by(date, time, station) %>% 
  mutate(sample_id = cur_group_id()) %>% 
  ungroup() %>% 
  relocate(sample_id)

```

### daily counts

Generate a 'seine' dataframe with

```{r}
# pivot to simplify summing Chinook counts (including 0's) for each sample unit
wide <-
  dat %>% 
  pivot_wider(
    # relies on 'sample_id' to distinguish among sample units, but retains environmental data
    id_cols = c(sample_id:time, water_temp),
    # columns for all fish spp (only CHN needed for this analysis)
    names_from = organism_code,
    # log number of for each seine (sample_id)
    values_from = count,
    # if no matches, enter '0'
    values_fill = 0,
    # sum counts if multiple matches
    values_fn = sum
  ) %>% 
  select(c(sample_id:water_temp, CHN))

# pivot back to get daily counts for each sample unit for CHN including 0's
# note that 'seine' includes only data for CHN...but should include all sample
# units (ie unique combos of site+date+time)
seine <- 
  wide %>% 
  pivot_longer(
  cols = "CHN",
  names_to = "CHN",
  values_to = "count")

### daily counts -----
# aggregate to daily totals bc freq >>1 seine per day
daily <- 
  seine %>%
  group_by(wy, wd) %>%
  summarise(
    daily_total = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

### cumulative percent -----
# calc cumulative percentage for each wy
daily_cum <- 
  daily %>%
  group_by(wy) %>%
  arrange(wd) %>%
  mutate(
    annual_total = sum(daily_total),
    cum_total = cumsum(daily_total),
    cum_percent = 100 * cum_total / annual_total
  ) %>% 
  ungroup() %>% 
  arrange(wy, wd)

daily_cum92 <-
  daily_cum %>% 
  filter(wy >= 1992)
```

## accumulation curves

Plot annual accumulation rates for juvenile Chinook as they enter the Delta

```{r}
# changed the colors where the older years are lighter and newer years are darker to see the phenological shifts in timing 
accumulation_curves <-
  ggplot(daily_cum,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       color = "water year") +
  theme_bw()

# outlier is 1977
accumulation_curves

# limit data to 1992+
accumulation_curves92 <-
  ggplot(daily_cum92,
       aes(x = wd,
           y = cum_percent,
           group = wy,
           color = wy)) +         
  geom_line() +
  geom_hline(yintercept = c(10, 50, 90),
             linetype = "dashed") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  labs(x = "Day of Water Year",
       y = "Cumulative % of Annual Total",
       title = "Juvenile Chinook Accumulation Curves",
       subtitle = "water years 1992-2024",
       color = "water year") +
  theme_bw()

accumulation_curves92
```

If there's a trend post-1992, it's hard to see in the accumulation curves.

## benchmarks

To examine the year-to-year differences more closely, I set (arbitrary) 10%, 50% and 90% accumulation benchmarks ("day_10", "day_50", and "day_90") for the migration timing, and compared these values across years.

```{r}
#| fig-cap: "Day of the Water Year for each Benchmark (10, 50, 90%)"

## ID benchmarks ----
# day of water year when the 10%, 50% and 90% of total annual numbers are reached

# using 10%, 50%, and 90% accumulation benchmarks (arbitrary)
benchmarks <- c(10, 50, 90)

# gen df of water year days when threshold reached for each year
percentile_days <- 
  daily_cum %>%
  group_by(wy) %>%
  summarise(
    day_10 = min(wd[cum_percent >= 10]),
    day_50 = min(wd[cum_percent >= 50]),
    day_90 = min(wd[cum_percent >= 90]),
    .groups = "drop"
  )

# pivot data to long format to facilitate graphics
pctl_days <-
  percentile_days %>% 
  pivot_longer(cols = c(day_10, day_50, day_90), 
               names_to = "percentile", 
               values_to = "wd") %>% 
  mutate(percentile = factor(percentile))

# same data without 1997
pctl_days_red <-
  pctl_days %>% 
  filter(wy != "1977")

pctl_days_92 <-
  pctl_days %>% 
  filter(wy >= 1992)

# figure shows the day of the water year for each benchmark (10, 50, 90%) for each year

benchmark_trends <-
  pctl_days %>% 
  ggplot(
    # color points according to threshold (10, 50, 90%)
    aes(x = wy, y = wd, color = percentile)
  ) +
  geom_point(size = 4) +
  scale_color_fish(discrete = TRUE,
                  option = "Oncorhynchus_tshawytscha",
                  alpha = 0.6) +
  geom_smooth(data = pctl_days_red,
              alpha = 0.2,
              # show linear trends
              method = "lm",
              se = TRUE) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "1977 excluded from trend analyses",
       x = "Water Year",
       y = "Day of the Water Year",
       color = "Percentile") +
  theme_bw()

benchmark_trends

# benchmark trends using data from 1992+
benchmark_trends92 <-
  pctl_days %>% 
  filter(wy >= 1992) %>% 
  ggplot(
    # color points according to threshold (10, 50, 90%)
    aes(x = wy, y = wd, color = percentile)
  ) +
  geom_point(size = 4) +
  scale_color_fish(discrete = TRUE,
                  option = "Oncorhynchus_tshawytscha",
                  alpha = 0.6) +
  geom_smooth(data = pctl_days_92,
              alpha = 0.2,
              # show linear trends
              method = "lm",
              se = TRUE) +
  labs(title = "Chinook Salmon Outmigration Timing",
       subtitle = "water years 1992-2024",
       x = "Water Year",
       y = "Day of the Water Year",
       color = "Percentile") +
  theme_bw()

benchmark_trends92
```

I found that, over time (1978-2025), all three thresholds were reached earlier (benchmark_trends). To explore this relationship, I modeled the timing for each benchmark (10, 50 and 90% of each year's total outmigrants) as a function of the water year.

```{r}
#| include: false

# reaching the 10% benchmark
m.day10.wy <- 
  brm(data = percentile_days, # use the percentile_days data
      # use a gaussian (normal) distribution
      family= gaussian,
      # specify the model
      day_10 ~ 1 + wy,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup= 1000, chains= 4, cores=4,
      # setting the "seed" determines which randomnumbers will get sampled
      # so that team gets the same results when running the model
      seed = 42,
      file = here("salmon/results/m.day10.wy"))

# reaching the 50% benchmark
m.day50.wy <- 
  brm(data = percentile_days, 
      # Choose a gaussian (normal) distribution
      family= gaussian,
      day_50 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/m.day50.wy"))

# reaching the 90% benchmark
m.day90.wy <- 
  brm(data = percentile_days, 
      family= gaussian,
      day_90 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      seed = 42,
      file = here("salmon/results/m.day90.wy"))
```

```{r}
summary(m.day10.wy)
summary(m.day50.wy)
summary(m.day90.wy)
```

```{r}
#| include: false

# run the model plots for m.d_10.wy, m.d_50.wy, m.d_90.wy
plot(m.day10.wy)
plot(m.day50.wy)
plot(m.day90.wy) # less certain than the others

# calc the prob that the slope is less than 0
m.day10.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
# virtually certain!

m.day50.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m.day90.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
```

### limited data

I ran those analyses again, limiting them to data from water year 1992 on.

```{r}
#| include: false

percentile_days92 <-
  percentile_days %>% 
  filter(wy >= 1992)

# reaching the 10% benchmark
m2.day10.wy <- 
  brm(data = percentile_days92, # use the percentile_days92 data
      # use a gaussian (normal) distribution
      family= gaussian,
      # specify the model
      day_10 ~ 1 + wy,
      # specify parameters for executing the Markov chains
      iter = 2000, warmup= 1000, chains= 4, cores=4,
      # setting the "seed" determines which randomnumbers will get sampled
      # so that team gets the same results when running the model
      seed = 42,
      file = here("salmon/results/m2.day10.wy"))

# reaching the 50% benchmark
m2.day50.wy <- 
  brm(data = percentile_days92, 
      # Choose a gaussian (normal) distribution
      family= gaussian,
      day_50 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 42,
      file = here("salmon/results/m2.day50.wy"))

# reaching the 90% benchmark
m2.day90.wy <- 
  brm(data = percentile_days92, 
      family= gaussian,
      day_90 ~ 1 + wy,
      iter = 2000, warmup = 1000, chains = 4, cores =4,
      seed = 42,
      file = here("salmon/results/m2.day90.wy"))

```

```{r}
summary(m2.day10.wy)
summary(m2.day50.wy)
summary(m2.day90.wy)
```

```{r}
plot(m2.day10.wy)
plot(m2.day50.wy)
plot(m2.day90.wy)
```

```{r}
# model results aren't encouraging, but let's calc the prob that the slope is less than 0 just for fun

m2.day10.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m2.day50.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())

m2.day90.wy %>%
  spread_draws(b_wy) %>%
  summarise(p_slope_lessthan_zero = sum(b_wy < 0) / n())
```

I ran three separate Bayesian linear models to look at the timing of juvenile salmon outmigration, represented by the day of the water year when those three benchmarks had been reached over the course of the time series. Each model used a Gaussian distribution with water year as the predictor. Our 10% and 50% models' credible intervals provide strong evidence that the slopes are different from zero, however, the upper limit of the 90% model is at 0 exactly which gives less certainty to this slope's difference from zero and an overall weaker trend.

-   For the 10% accumulation model, there was an estimated slope of -0.69 days per year (probability the slope is \<0 = 0.999)

-   For the 50% accumulation model, there was an estimated slope of -0.70 days per year (probability the slope is \<0 = 1.000)

-   For the 90% accumulation model, there was an estimated slope -0.33 days per year (probability the slope is \<0 = 0.955)

## water year type

I also attempted to determine if there was an effect of water year type on the timing of Chinook outmigration. I tested several models that included both Water Year and Water Year Type as explanatory variables.

```{r}
## year type ----
# first, add water year type to the percentile_days df
wy_type <- 
  read_excel(here("salmon/data/raw/cdec-water-year-type-jun-2025.xlsx")) %>% 
  mutate(wy = WY,
         wyt = factor(WYT),
         watershed = factor(Area),
         .keep = "unused")

# note that I'm using the df withOUT 1977
pctl_days_wyt <-
  wy_type %>% 
  # select Sac watershed, excludes SJR watershed
  filter(watershed == "Sacramento Valley") %>% 
  # merge w simple df, no 1977 w problem data 
  right_join(pctl_days_red,
             by = "wy") %>% 
  dplyr::select(-watershed) %>% 
  # specify order of the wyt levels
  mutate(wyt = fct_relevel(wyt, "W", "AN", "BN", "D", "C"),
         # center wy to facilitate model convergence
         wy_centered = wy - 1978) %>% 
  relocate(wy_centered, .after = wy)

write_rds(pctl_days_wyt, 
          here("salmon/data/processed/pctl_days_wyt.rds"))

### wy boxplot -----
pctl_days_wyt %>% 
  filter(percentile == "day_10",
         wy <= 2024) %>% 
  ggplot(aes(x = wyt, y = wd, group = wyt)) +
  geom_boxplot() +
  labs(title = "Timing of 10% Annual Accumulation Threshold",
       x = "Water Year Type",
       y = "Day of the Water Year") +
  theme_bw()

### m1 ----
# run wy + wyt model with the centered variable, 1977 & 2025 not included
m1.day10.wy.wyt <- 
  brm(data = pctl_days_wyt %>% 
        filter(wy <= 2024),
      family = gaussian,
      wd ~ 0 + wy_centered + wyt, 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, 
      seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("salmon/results/m2.day10.wy.wyt")
  )

summary(m1.day10.wy.wyt) 
plot(m1.day10.wy.wyt)
# Rhat is perfect
# water year effect looks solid (albeit kinda tight on 0)
# posterior distributions look great (uniform)
# mcmc look great too
# estimates for year type are wonky--reflect the data! 
# is there an interaction effect?
# should also include some semi-informed priors

### m2 ----
# run wy * wyt model with the centered variable, 1977 not included
m2.day10.wy.wyt <- 
  brm(data = pctl_days_wyt %>% 
        filter(wy <= 2024),
      family = gaussian,
      wd ~ 0 + wy_centered * wyt, 
      iter = 10000, warmup = 1000, chains = 4, cores = 4, 
      seed = 42,
      # save fitted model (this one w the wy centered)
      file = here("salmon/results/m3.day10.wy.wyt")
  )

summary(m2.day10.wy.wyt)
plot(m2.day10.wy.wyt)
```

These results are not so solid: All four of the interaction terms span 0.

Posterior predictive check...see if the model's "noise" (sigma) matches the actual spread of your data. These look good for both models.

```{r}
# posterior predictive check
pp_check(m1.day10.wy.wyt)
pp_check(m2.day10.wy.wyt)
```

Because the model simulations (blue lines) approximate the actual data fairly well (they aren't wider or narrower), this sigma seems to be a good estimation of the error.
