---
title: "hydrology"
author: "Pete Nelson"
editor: visual
bibliography: references.bib
---

## introduction

Using the approaches described in [@wang2024][@archfield2014], I've tried to characterize the flow regime for the Sacramento River as it enters the Delta. These include the Magnificent 7 [@archfield2014a; @helsel2020] as well as....

## setup

```{r}
#| label: libraries

library(here)
library(tidyverse)
library(Rbeast)
library(hydrostats)
library(envPred)
library(readxl)
library(EflowStats)
library(dataRetrieval)
library(slider)

```

## data

The data used here ('dayflow.xlsx') were downloaded from <https://data.cnra.ca.gov/dataset/dayflow> using the 'Export' link (select 'CSV'), and compiled from 1970-1983, 1984-1996, 1997-2023, 2024, and 2025 (completed 16 April 2026). Meta data are available here: <https://data.cnra.ca.gov/dataset/06ee2016-b138-47d7-9e85-f46fae674536/resource/776b90ca-673e-4b56-8cf3-ec26792708c3/download/current-dayflow-documentation.pdf> These data include flow measurements from multiple systems; I've selected the 'SAC' data specifically below. The SAC data for the Sacramento River are from USGS AVM monitoring and stream gage data, DWR Station Numbers B9-1840 (Freeport) and A0-2100 (Sacramento) and USGS Station Numbers 11-4467.50 and 11-4475.00.

```{r}
#| label: dayflow


# dayflow
flow <- read_excel(here("data/raw/dayflow.xlsx"))

# convert to standard data.frame and fix the date class
flow_df <- as.data.frame(flow[, c("Date", "SAC")])
flow_df$Date <- as.Date(flow_df$Date)

```

## magnificent 7

The 'Magnificent 7' are seven fundamental daily streamflow statistics (FDSS) used to characterize natural flow regimes and streamflow time series (see [@archfield2014], and references therein). These are:

1.  Mean flow \[lam1\]

2.  Coefficient of Variation (flow variability) \[tau2\]

3.  Skewness (asymmetry of the flow distribution) \[tau3\]

4.  Kurtosis (peakedness of the distribution) \[tau4\]

5.  Autoregressive Lag-One Correlation Coefficientn (AR(1)) (persistence or duration) \[ar1\]

6.  Amplitude (of seasonal signal) \[amplitude\]

7.  Phase/Timing (of seasonal signal) \[phase\]

```{r}
#| label: mag-7

# run the function
m7 <- calc_magnifSeven(flow_df)

# view results
print(m7)

```

### magnitude & variability

The first four indices (`lam1` through `tau4`) use "L-moments," more robust versions of standard statistics like Mean and Skewness.

- **`lam1` (22,690.48):** This is the *mean daily flow* (cfs) for the lower Sacramento River.

- **`tau2` (0.38):** This is the *L-Coefficient of Variation*. A value of 0.38 indicates moderate variability. It’s not a steady canal, but it’s also not a "flashy" desert stream (\> 0.6).

- **`tau3` (0.43):** This is the *L-Skewness*. A value of 0.43 is quite high. It confirms that the Sacramento’s "average" is driven by a few massive flood events rather than consistent daily flow.

- **`tau4` (0.24):** This is the *L-Kurtosis*. It measures the "fatness" of the tails. This suggests that extreme outliers (huge floods) happen more often than they would in a standard bell curve.

### persistence & seasonality

- **`ar1` (0.989):** This is a measure of *autocorrelation*. A value of 0.989 is extremely high, but typical of large rivers or regulated systems where flow does not change abruptly.

- **`amplitude` (0.630):** This is the *seasonal signal strength*. With a fairly high value—closer to 1 than 0--, this is indicative of a strong seasonal cycle (i.e., winter rain and spring snowmelt).

- **`phase` (61.358):** This is the *seasonal timing* (in the numerical day of the water year). A value of ca. 61 indicates that the peak of the seasonal "wave" occurs roughly 61 days into the water year (November 30).

These results show a **large, highly predictable, highly persistent river** that is prone to **large, infrequent flood spikes** (high skewness).

When does the peak seasonal flow occur for each year in our time series?

```{r}
#| label: peak-flow

peak_flow <-
  flow_df %>% 
  # add the water year (wy)
  mutate(wy = get_waterYear(Date)) %>% 
  group_by(wy) %>% 
  # find the day with the max flow
  filter(SAC == max(SAC, na.rm = T)) %>% 
  # calc day of water year when peak flow occurred
  mutate(peak_day = as.numeric(Date - as.Date(paste0(wy - 1, "-10-01"))) + 1) %>%
  select(wy, Date, peak_day, SAC)

peak_flow %>% 
  ggplot(aes(x = wy, y = peak_day)) +
  geom_point() +
  geom_smooth(method = "lm", se = T, color = "blue") +
  theme_bw()

m1.peak <-
  brm(data = peak_flow,
    peak_day ~ wy,
    family = gaussian,
    prior(normal(0, 25)),
    iter = 2000,
    warmup = 1000,
    chains = 4,
    cores = 4,
    seed = 42
    )

print(m1.peak)
plot(m1.peak)

```

## environmental predictability

Determine the predictability [@colwell1974] of flow in the Sacramento River using dayflow data (wy 1956-2026). Most useful as a tool for comparing different systems (e.g., Sac R vs SJR), I'm including this as a place holder for future work comparing (perhaps) multiple spawning tributaries.

```{r}
#| label: flow-table

sac_envpred <-
  flow_df %>% 
  mutate(Q = SAC, .keep = "unused") %>% 
  Colwells(.,
           s = 11)

sac_envpred$flow.table

```

Let's use library envPred to calculate environmental predictability and few other metrics.

```{r}
#| label: env-pred

# scientific notation = FALSE
options(scipen = 999)

# select the metrics I find useful
metrics_i_want <- c("n_yrs", "n_days", "n_na",
                    "unbounded_seasonality", "bounded_seasonality",
                    "env_col",
                    "colwell_p", "colwell_c", "colwell_m")

sac_envpred2 <-
  env_stats(time_series = flow_df$SAC,
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

# log-transform the flow data
sac_envpred3 <-
  env_stats(time_series = log(flow_df$SAC),
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

# you can ignore the warning about the ts starting/ending diff times of the year

```

## BEAST

*Bayesian Estimator of Abrupt change, Seasonality, and Trend* (BEAST) is a Bayesian model averaging algorithm, used to decompose time series into individual components, including abrupt changes, trends, and periodic or seasonal variations .

```{r}
#| label: time-series

sac_beast <-
  beast(flow_df$SAC, 
        deltat = 1/365.25, 
        start = 1955,
        period = 1,
        mcmc.samples = 10000, # increased for better convergence
        max.trend.knots = 30, # default (10) had it maxing out
        max.season.knots = 20) # allow more seasonal shifts

print(sac_beast)
saveRDS(sac_beast, here("salmon/results/sac_beast.rds"))

sac_beast <-
  read_rds(here("salmon/results/sac_beast.rds"))

plot(sac_beast)

pdf("salmon/plots/SacR_ts.pdf", width = 11, height = 8.5)
plot(sac_beast)

# png("salmon/plots/SacR_ts.png", width = 2400, height = 1800, res = 300)

```

## first flush

Identifying peak flow events.

```{r}

peaks <- flow_df %>% 
  arrange(Date) %>% 
  mutate(
    # calc 3-day moving average
    ma3 = slide_dbl(SAC, mean, .before = 2, .complete = T),
    # daily change in the 3-day average
    ma3_diff = ma3 - lag(ma3),
    # define condition: change >= 2% of the change
    is_condition_met = ifelse(ma3_diff > 0, TRUE, FALSE),
    # rising flows for at least 4 days
    streak = slide_dbl(is_condition_met, sum, .before = 3, .complete = T),
    is_event_start = streak == 4
  )

peaks %>% 
  ggplot(aes(x = Date, y = SAC)) +
  geom_line() +
  theme_bw()

peaks %>% 
  filter(year(Date) >= 1970 & year(Date) < 1980) %>% 
  ggplot(aes(x = Date, y = SAC)) +
  geom_line(color = "blue") +
  theme_bw()

peaks %>% 
  filter(Date >= "2024-10-01" & Date <= "2025-09-30") %>% 
  ggplot(aes(x = Date, y = SAC)) +
  geom_line(color = "gray", alpha = 0.5) +
  geom_line(aes(color = is_condition_met, group = 1), linewidth = 1) +
  scale_color_manual(
    values = c("TRUE" = "blue", "FALSE" = "green"),
    labels = c("TRUE" = "Rising", "FALSE" = "Descending"),
    name = "flow phase"
  ) +
  theme_bw()


# Assuming your dataframe is 'df' with columns 'Date' and 'SAC'
df_plotted <- flow_df %>%
  arrange(Date) %>%
  mutate(
    # Rising logic: >10% over 3 days
    pct_change_3d = (SAC - lag(SAC, 3)) / lag(SAC, 3),
    # Descending logic: >2% decrease (<-0.02) over 1 day
    pct_change_1d = (SAC - lag(SAC, 1)) / lag(SAC, 1),
    
    # Assign Limb Status
    limb_type = case_when(
      pct_change_3d > 0.10 ~ "Rising",
      pct_change_1d < -0.02 ~ "Descending",
      TRUE ~ "Stable"
    )
  )

# Plotting with segments to color the line itself
df_plotted %>% 
  filter(Date >= "2024-10-01") %>% 
ggplot(aes(x = Date, y = SAC, color = limb_type, group = 1)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Rising" = "red", 
                                "Descending" = "blue", 
                                "Stable" = "grey")) +
  theme_minimal() +
  labs(title = "Sacramento River Hydrograph Limbs",
       y = "Discharge (cfs)",
       color = "Limb Status")


```
