---
title: "hydrology"
author: "Pete Nelson"
editor: visual
bibliography: references.bib
---

## introduction

Using the approaches described in [@wang2024][@archfield2014], I've tried to characterize the flow regime for the Sacramento River as it enters the Delta. These include the Magnificent 7 [@archfield2014a; @helsel2020] as well as....

## setup

```{r}
#| label: libraries

library(Rbeast)
library(hydrostats)
library(envPred)

```

## data

```{r}
#| label: dayflow


# dayflow
flow <- read_excel(here("data/raw/dayflow.xlsx"))

# convert to standard data.frame and fix the date class
flow_df <- as.data.frame(flow[, c("Date", "SAC")])
flow_df$Date <- as.Date(flow_df$Date)

```

## magnificent 7

```{r}
#| label: mag-7

# run the function
m7 <- calc_magnifSeven(flow_df)

# view results
print(m7)

```

### magnitude & variability

The first four indices (`lam1` through `tau4`) use "L-moments," more robust versions of standard statistics like Mean and Skewness.

- **`lam1` (22,690.48):** This is the *mean daily flow* (cfs) at this location.

- **`tau2` (0.38):** This is the *L-Coefficient of Variation*. A value of 0.38 indicates moderate variability. It’s not a steady canal, but it’s also not a "flashy" desert stream (\> 0.6).

- **`tau3` (0.43):** This is the *L-Skewness*. A value of 0.43 is quite high. It confirms that the Sacramento’s "average" is driven by a few massive flood events rather than consistent daily flow.

- **`tau4` (0.24):** This is the *L-Kurtosis*. It measures the "fatness" of the tails. This suggests that extreme outliers (huge floods) happen more often than they would in a standard bell curve.

### persistence & seasonality

- **`ar1` (0.989):** This is a measure of *autocorrelation*. A value of 0.989 is extremely high, but typical of large rivers or regulated systems where flow does not change abruptly.

- **`amplitude` (0.630):** This is the *seasonal signal strength*. With a fairly high value—closer to 1 than 0--, this is indicative of a strong seasonal cycle (i.e., winter rain and spring snowmelt).

- **`phase` (61.358):** This is the *seasonal timing* (in days). A value of ca. 61 indicates that the peak of the seasonal "wave" occurs roughly 61 days into the water year (early December).

These results show a **large, highly predictable, highly persistent river** that is prone to **large, infrequent flood spikes** (high skewness).

## environmental predictability

Determine the predictability [@colwell1974] of flow in the Sacramento River using dayflow data (wy 1956-2026). Most useful as a tool for comparing different systems (e.g., Sac R vs SJR), I'm including this as a place holder for future work comparing (perhaps) multiple spawning tributaries.

```{r}
#| label: flow-table


library(hydrostats)

sac_envpred <-
  flow_df %>% 
  mutate(Q = SAC, .keep = "unused") %>% 
  Colwells(.,
           s = 11)

sac_envpred$flow.table

```

Let's use library envPred to calculate environmental predictability and few other metrics.

```{r}
#| label: env-pred

library(envPred)
# scientific notation = FALSE
options(scipen = 999)

# some metrics I don't find useful
metrics_i_want <- c("n_yrs", "n_days", "n_na",
                    "unbounded_seasonality", "bounded_seasonality",
                    "env_col",
                    "colwell_p", "colwell_c", "colwell_m")

sac_envpred2 <-
  env_stats(time_series = flow_df$SAC,
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

# log-transform the flow data
sac_envpred3 <-
  env_stats(time_series = log(flow_df$SAC),
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

```

## BEAST

*Bayesian Estimator of Abrupt change, Seasonality, and Trend* (BEAST) is a Bayesian model averaging algorithm, used to decompose time series into individual components, including abrupt changes, trends, and periodic or seasonal variations .

```{r}
#| label: time-series

sac_beast <-
  beast(flow_df$SAC, 
        deltat = 1/365.25, 
        start = 1955,
        period = 1,
        mcmc.samples = 10000, # increased for better convergence
        max.trend.knots = 30, # default (10) had it maxing out
        max.season.knots = 20) # allow more seasonal shifts

print(sac_beast)
saveRDS(sac_beast, here("salmon/results/sac_beast.rds"))

plot(sac_beast)

pdf(here("salmon/plots/SacR_ts.pdf"), width = 11, height = 8.5)
plot(sac_beast)

png(here("salmon/plots/SacR_ts.png"), width = 2400, height = 1800, res = 300)
plot(sac_beast)

dev.off()

```
