---
title: "hydrology"
author: "Pete Nelson"
editor: visual
bibliography: references.bib
---

## introduction

Using the approaches described in [@wang2024][@archfield2014], I've tried to characterize the flow regime for the Sacramento River as it enters the Delta. These include the Magnificent 7 [@archfield2014a; @helsel2020] as well as....

## setup

```{r}
#| label: libraries

library(Rbeast)
library(hydrostats)
library(envPred)

```

## data

```{r}
#| label: dayflow


# dayflow
flow <- read_excel(here("data/raw/dayflow.xlsx"))

# convert to standard data.frame and fix the date class
flow_df <- as.data.frame(flow[, c("Date", "SAC")])
flow_df$Date <- as.Date(flow_df$Date)

```

## magnificent 7

```{r}
#| label: mag-7

# run the function
m7 <- calc_magnifSeven(flow_df)

# view results
print(m7)

```

**1. Magnitude and Variability (The "L-Moments")**

The first four indices (`lam1` through `tau4`) use "L-moments," which are more robust versions of standard statistics like Mean and Skewness.

-   **`lam1` (22,690.48):** This is the **Mean Daily Flow**. Your average flow at this location is roughly 22,691 cfs.

-   **`tau2` (0.38):** This is **L-Coefficient of Variation**. A value of 0.38 indicates moderate variability. It’s not a steady canal, but it’s also not a "flashy" desert stream (which would be \> 0.6).

-   **`tau3` (0.43):** This is **L-Skewness**. A value of 0.43 is quite high. It confirms that the Sacramento’s "average" is driven by a few massive flood events rather than consistent daily flow.

-   **`tau4` (0.24):** This is **L-Kurtosis**. It measures the "fatness" of the tails. This suggests that extreme outliers (huge floods) happen more often than they would in a standard bell curve.

------------------------------------------------------------------------

**2. Persistence and Seasonality**

-   **`ar1` (0.989):** This is **Autocorrelation**. A value of 0.989 is **extremely high**.

    -   **Interpretation:** Today’s flow is 99% predicted by yesterday’s flow. This is typical of large, snow-fed rivers or regulated systems where flow doesn't change abruptly.

-   **`amplitude` (0.630):** This is the **Seasonal Signal Strength**.

    -   **Interpretation:** Since it’s closer to 1 than 0, the river has a strong, predictable seasonal cycle (likely the winter rain/spring snowmelt pattern).

-   **`phase` (61.358):** This is the **Seasonal Timing** (in days).

    -   **Interpretation:** A value of \~61 indicates that the peak of the seasonal "wave" occurs roughly 61 days into the water year (which starts Oct 1). This places the peak in **early December**.

------------------------------------------------------------------------

**Is it only useful for comparisons?**

It is **mostly** used for comparisons, but it serves two distinct purposes:

1.  **Hydrological Classification:** You can use these 7 numbers to "cluster" rivers. For example, rivers with `ar1` \> 0.9 and `amplitude` \> 0.5 are usually "Snowmelt/Large Basin" types.

2.  **Model Validation:** If you are building a computer model to simulate the Sacramento River, your model's output should produce these same 7 numbers. If your model gives a `tau3` of 0.1, you know your model is failing to capture the big flood peaks.

**⚓ Key Takeaway**

Your results show a **large, highly predictable, highly persistent river** that is prone to **large, infrequent flood spikes** (high skewness).

## environmental predictability

Determine the predictability [@colwell1974] of flow in the Sacramento River using dayflow data (wy 1956-2026). Most useful as a tool for comparing different systems (e.g., Sac R vs SJR), I'm including this as a place holder for future work comparing (perhaps) multiple spawning tributaries.

```{r}
#| label: flow-table


library(hydrostats)

sac_envpred <-
  flow_df %>% 
  mutate(Q = SAC, .keep = "unused") %>% 
  Colwells(.,
           s = 11)

sac_envpred$flow.table

```

Let's use library envPred to calculate environmental predictability and few other metrics.

```{r}
#| label: env-pred

library(envPred)
# scientific notation = FALSE
options(scipen = 999)

# some metrics I don't find useful
metrics_i_want <- c("n_yrs", "n_days", "n_na",
                    "unbounded_seasonality", "bounded_seasonality",
                    "env_col",
                    "colwell_p", "colwell_c", "colwell_m")

sac_envpred2 <-
  env_stats(time_series = flow_df$SAC,
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

# log-transform the flow data
sac_envpred3 <-
  env_stats(time_series = log(flow_df$SAC),
          dates = flow_df$Date,
          n_states = 11,
          delta = 1,
          # assume the ts is evenly distributed, so use 'spectrum'
          noise_method = "spectrum") %>% 
  as.list() %>% 
  as_tibble() %>% 
  # pivot the output so it's easier to read
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(value = round(value, 4)) %>% 
  filter(metric %in% metrics_i_want) %>% 
  print()

```

## BEAST

Bayesian Estimator of Abrupt change, Seasonality, and Trend (BEAST is a fast, generic Bayesian model averaging algorithm to decompose time series or 1D sequential data into individual components, such as abrupt changes, trends, and periodic or seasonal variations .

```{r}

sac_beast <-
  beast(flow_df$SAC)

print(sac_beast)
plot(sac_beast)


```
