---
title: "beach seine sites"
author: "Pete Nelson"
editor: visual
---

# tidy beach seine data

The beach seine data, as downloaded from EDI and using their code, has some issues with variable type and naming. This code addresses these issues and saves the beach seine data as `raw.rds`.

```{r}
#| label: tidy-data
#| message: false
#| warning: false

library(janitor)
library(here)
library(chron)
library(gridExtra)
library(sf)
library(tigris)
library(nhdplusTools)
library(tidyverse)

raw <- 
  read_rds(here("salmon/data/raw/dt3.rds")) %>% 
  # provide good names
  clean_names() %>% 
  # chng region & station to chr (factors interfere w regularization later)
  # fix date & time variables, originally as chr
  mutate(region = as.character(region_code),
         station = as.character(station_code),
         date = mdy(sample_date),
         time = times(sample_time),
         .keep = "unused") %>% 
  # generate a sample unit (unique seine event) code
  group_by(station, date, time) %>% 
  mutate(id_code = cur_group_id()) %>% 
  relocate(id_code, station, location, date, time) %>% 
  ungroup() %>% 
  write_rds(., here("salmon/data/raw/raw.rds"))

```

# check data quality

Run some quick checks on methodological and environmental data. The high values for specific_conductance appear to be from SF Bay (ie where you'd expect to find high values).

```{r}
#| label: data-qc
#| message: false
#| warning: false

# estimated volume of the seine effort
vol <- ggplot(raw,
              aes(volume)) +
  geom_histogram()

# water temperature (deg C)
temperature <-
  ggplot(raw,
         aes(water_temp)) +
  geom_histogram()

# turbidity
turb <-
  ggplot(raw,
         aes(turbidity)) +
  geom_histogram()

# specific conductance
cond <-
  ggplot(raw,
         aes(specific_conductance)) +
  geom_histogram()

# arrange graphs
grid.arrange(vol, temperature, turb, cond, ncol = 2, nrow = 2)

rm(vol, temperature, turb, cond)

```

# effort

Regularize the time series, and identify the months for which each site was sampled and *not* sampled. Add descriptions for regions. The file `coverage.rds` is a regularized time series showing which station/year/months were and weren't sampled.

```{r}
#| label: effort

# create monthly coverage indicator and retain region
# this is an irregular time series-no rows when sites weren't sampled
coverage_irreg <- raw %>%
  distinct(
    station, year = year(date), month = month(date), region) %>%
  # 'tag' months sampled (thus far, all of them)
  mutate(sampled = 1)

# create tibble of station and region for later use
# 1= Lower Sacramento River, 2= North Delta, 3= Central Delta, 4= South Delta, 
# 5= San Joaquin River, 6= Bay Area, 7= Sacramento River
sta_reg <-
  coverage_irreg %>%
  distinct(station, region) %>%
  left_join(
    tibble(
      region = factor(1:7),
      region_label = factor(c("Lower Sacramento River", "North Delta", 
                              "Central Delta", "South Delta",
                              "San Joaquin River", 
                              "San Francisco & San Pablo Bays", 
                              "Sacramento River"))
    ),
    by = "region"
  ) %>% 
  write_rds(., here("salmon/data/processed/sta_reg.rds"))

# regularize ts
coverage <-
  coverage_irreg %>% 
  # regularize the time series w complete()
    complete(
        station,
        # add missing years
        year = min(year):max(year),
        # add missing months
        month = 1:12,
        # enter '0' for station/year/months that weren't sampled
        fill = list(sampled = 0)
    ) %>% 
  # add region code for unsampled station/year/months (lost w regularization)
    rows_patch(sta_reg %>% 
                 dplyr::select(station, region), by = "station") %>%
  # add region_label
    left_join(sta_reg %>% distinct(region, region_label), by = "region") %>% 
    mutate(
      month_label = factor(month, levels = 1:12, labels = month.abb),
      region_label = fct_relevel(
        region_label, "Sacramento River", "Lower Sacramento River", 
        "North Delta", "Central Delta",
        "San Joaquin River", "South Delta", "San Francisco & San Pablo Bays")
    ) %>% 
  write_rds(., here("salmon/data/processed/coverage.rds"))

```

# plot

## effort

Plot monthly coverage for all beach sites.

```{r}
#| label: heat-maps


# generate calendar-style heat map 
# months run horizontally, years vertically

# too many sites to adequately display the information, so divide them between
# northerly sites and southerly ones

# North Delta effort
coverage %>% 
  filter(region_label == "North Delta",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "midnightblue"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: North Delta"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
n_delta <- c("AM001S", "MS001N", "SR014W", "SR017E", 
             "SR024E", "SR043W", "SR049E", "SR060E")

# Central Delta effort
coverage %>% 
  filter(region_label == "Central Delta",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "darkgreen"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: Central Delta"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
cen_delta <- c("DS002S", "GS010E", "LP003E", "MK004W", 
               "SF014E", "SJ001S", 
               "SJ005N", 
               "TM001N", "XC001N")

# South Delta effort
coverage %>% 
  filter(region_label == "South Delta",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "darkred"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: South Delta"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
s_delta <- c("MR010W", "SJ041N")

# San Francisco & San Pablo bay effort
coverage %>% 
  filter(region_label == "San Francisco & San Pablo Bays",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "darkgray"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: San Francisco & San Pablo Bays"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
# none of these sites qualified for sufficient coverage; SJ056E is good post-1997; SJ058W & SJ063W are marginally acceptable post-1997; I looked at but discarded SJ068W, SJ074W, SJ077E, SJ083W

# San Joaquin River effort
coverage %>% 
  filter(region_label == "San Joaquin River",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "blue"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: San Joaquin River"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
# none of these sites qualified for sufficient coverage

# Sacramento River effort
coverage %>% 
  filter(region_label == "Sacramento River",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "orange"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: Sacramento River"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
# none of these sites qualified for sufficient coverage

# Lower Sacramento River effort
coverage %>% 
  filter(region_label == "Lower Sacramento River",
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, fill = factor(sampled))) +
  geom_tile(color = NA) +
  facet_wrap(~station) +
  scale_fill_manual(values = c("0" = "grey95", "1" = "chartreuse"), guide = "none") +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage: Lower Sacramento River"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 7)
  )
l_sac <- c("SR071E", "SR080E", "SR090W", "SR094E", 
           "SR138E")


```

## site selection

I basically eye-balled the figures above to select the following sites as judged minimally adequate or better for their temporal coverage.

```{r}
#| label: site-selection

# 'eye-balled' the figures to identify the 'good' sites, those with better
# annual and seasonal coverage (some still sparse or late start/early end):

# here are all the DJFMP sites w RKM
stations_with_rkm <-
  read_rds(here("salmon/data/processed/stations_with_rkm.rds"))

# this is a pretty conservative selection
sites <- 
  tibble(station = c(cen_delta, l_sac, n_delta, s_delta)) %>% 
  left_join(stations_with_rkm, by = "station") %>% 
  arrange(., region) %>% 
  # river_km is approx dist from confluence
  # RKI is the kms to the Golden Gate, following thalweg
  mutate(RKI = river_km + 86) %>% 
  write_rds(., here("salmon/data/processed/sites.rds"))

# sites selected for complete-ish coverage post-1998; here are MORE 
# sites but suitable for use with a restricted timeframe (1998+)
# sites w good data post-2000 based on looking at heat maps
new_sites <-
  # north delta
  c("AM001S", "MS001N", "SR014W", "SR017E", "SR024E", "SR043W", "SR049E", "SR060E", "SS011N",
    # central delta
    "DS002S", "GS010E", "LP003E", "MK004W", "SF014E", "SJ001S", "SJ005N", "TM001N", "XC001N",
    # south delta
    "MR010W", "SJ041N", "SJ032S", "SJ051E",
    # sf and san pablo bays
    "SA001M", "SA004W", "SA007E", "SA008W", "SA009E", "SA010W", "SP000W", "SP001W", "SP003E",
    # san joaquin
    "SJ056E", "SJ058W", "SJ063W",
    # lower sac r
    "SR071E", "SR080E", "SR090W", "SR094E", "SR138E"
  )

sites_ltd <-
  tibble(station = new_sites) %>% 
  left_join(stations_with_rkm, by = "station") %>% 
  arrange(., region) %>% 
  mutate(RKI = river_km + 86) %>% 
  write_rds(., here("salmon/data/processed/sites_ltd.rds"))

# what does effort look like by region?
coverage %>% 
  # select stations from the 'sites' tibble, those w consistent effort Jan-Mar
  filter(station %in% sites$station,
         year >= 1977) %>% 
  mutate(month_label = 
           factor(month, levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(month_label, year, 
             # sampled: 0/1  (no/yes)
             fill = factor(sampled))) +
  geom_tile(color = NA) +
  # consider regions individually
  facet_wrap(~region_label) +
  labs(
    x = "Month",
    y = "Year",
    title = "Monthly Sampling Coverage by Region"
  ) +
  theme_bw()

# central delta (example)
coverage %>% 
    filter(station %in% sites$station,
           year >= 1977, region_label == "Central Delta") %>% 
    mutate(month_label = 
               factor(month, levels = 1:12, labels = month.abb)) %>% 
    ggplot(aes(month_label, year, 
               fill = factor(sampled))) +
    geom_tile(color = NA) +
    scale_fill_manual(
        name = "Sampled?",
        values = c("0" = "grey85", "1" = "steelblue"),
        labels = c("0" = "No", "1" = "Yes")
    ) +
    labs(
        x = "Month",
        y = "Year",
        title = "Monthly Beach Seine Effort",
        subtitle = "Central Delta sites (selected)"
    ) +
    theme_bw() +
  theme(legend.position = c(0.95, 0.13),
        legend.justification = c("right", "bottom"))


```

Jon Speegle sent me some sampling effort data from 2000 on. Using this, I determined which sites had been sampled all 27 water years across a median of 11+ months. A list of these sites is contained in site_sampling.csv.

# map sites

The first flush working group has adopted specific regions to explore spatial variation in first flush analyses. Here, I'm sorting selected beach seine sites into these working regional polygons and mapping the results.

```{r}
#| label: map-sites

# import Dave's regions shapefile
sf_regions <- 
  read_sf(here("data/processed/spatial/delta_regions.shp")) %>% 
  clean_names()

# create stations dataframe w lat lon
stations <-
  sites %>% 
  left_join(
    # retrieve DJFMP station metadata
    read_rds(here("salmon/data/raw/dt5.rds")),
            join_by(station == StationCode)) %>% 
  mutate(lat = Latitude,
         lon = Longitude,
         .keep = "unused") %>% 
  dplyr::select(-MethodCode) %>% 
  clean_names()

# convert stations to an sf object
stations_sf <-
  st_as_sf(stations,
           coords = c("lon", "lat"),
           # WGS84 lon/lat
           crs = 4326) %>% 
  # transform station points (NAD83/UTM zone 10N)
  st_transform(., st_crs(sf_regions)) %>% 
  # converts coordinates to recognizable lon/lat
  st_transform(4326) %>% 
  # add lon & lat, separate from geometry
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  ) %>% 
  # add first flush regions
  mutate(ff_region =
           factor(
             case_when(
             lat > 38.5   ~ "SAC",
             lon > -121.6 ~ "SJR",
             lon < -122.0 ~ "Bay",
         .default = as.character(region_label))
           )
  )
    
# filter regions to retain first flush priorities
stations_sf <-
  stations_sf %>% 
  filter(ff_region %in% c("SAC", "SJR", "Bay")) %>% 
  dplyr::select(station, region_label, location, ff_region) %>% 
  mutate(region_label = fct_relevel(region_label,
                                    "Lower Sacramento River", "North Delta",
                                    "Central Delta", "South Delta"))

# map regions, including sites outside ff region
ggplot() +
  geom_sf(data = sf_regions, fill = "grey95", color = "grey60") +
  geom_sf(
    data = stations_sf,
    aes(color = region_label),
    size = 2
  ) +
  labs(
    title = "Stations Colored by Region",
    color = "Region"
  ) +
  theme_bw()


```

Add streamlines and the estuary's outline.

```{r}
#| label: better-map


# enable map data caching
options(tigris_use_cache = TRUE)

# --- Bay/Estuary outlines ---
# pull area water for the relevant counties (Contra Costa, Marin, Solano, etc.)
bay_counties <- c("Contra Costa", "Marin", "Solano", "Napa", 
                  "Sonoma", "San Francisco", "Alameda")

bay_water <- map_dfr(bay_counties, 
                     ~ area_water("CA", county = .x, class = "sf")) %>%
  st_transform(st_crs(sf_regions))  # match existing CRS

bay_water_filtered <- bay_water %>%
  filter(AWATER > 10000000)  # units: square meters (10000000=10km^2)
# could also filter by FULLNAME

# --- NHD Streamlines ---
# define a bounding box around study area
bbox <- st_bbox(
  c(xmin = -122.6, ymin = 37.7, xmax = -121.2, ymax = 39.3),
  crs = st_crs(4326)
) %>% st_as_sfc()

# get NHD flowlines within the area, save!
flowlines <- get_nhdplus(AOI = bbox, realization = "flowline")
write_rds(flowlines, here("salmon/data/processed/flowlines"))

# filter to meaningful streams (avoids tiny ditches cluttering the map)
flowlines_filtered3 <- flowlines %>%
  filter(streamorde >= 3)

# --- updated plot ---
beach_seine_stations <-
  ggplot() +
  geom_sf(data = sf_regions, fill = "grey95", color = "grey60") +
  geom_sf(data = bay_water_filtered, fill = "lightsteelblue2", color = "steelblue3", linewidth = 0.3) +
  geom_sf(data = flowlines_filtered3, color = "steelblue4", linewidth = 0.4, alpha = 0.7) +
  geom_sf(data = stations_sf, aes(color = region_label), size = 2) +
  # Place coord_sf LAST and define the CRS inside it
  coord_sf(
    xlim = c(-122.6, -121.2), 
    ylim = c(37.7, 39.3), 
    crs = st_crs(4326), 
    default_crs = st_crs(4326),
    expand = FALSE
  ) +
  labs(title = "Selected Beach Seine Stations", color = "Region") +
  theme_bw()

ggsave("beach_seine_stations.png", 
       path = here("salmon/plots"),
       plot = beach_seine_stations, 
       width = 8, height = 8, 
       dpi = 300)



```

Add station labels and city anchors using ggrepel

```{r}
#| label: refine-map


library(ggrepel)

# reference cities for geographic context

cities <- data.frame(
  name = c("Colusa", "Sacramento", "Rio Vista", "Stockton", "Benicia", "San Francisco"),
  lon = c(-122.0094, -121.4789, -121.6926, -121.2908, -122.1539, -122.4194),
  lat = c(39.2143, 38.5758, 38.1606, 37.9577, 38.0521, 37.7749)
)

# had "Knights Landing" -121.7183, 38.7996 but there's a KL station

cities_sf <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# build combined labels dataframe
station_labels <- stations %>%
  select(label = location, lon, lat) %>%
  mutate(type = "station")

city_labels <- cities %>%
  select(label = name, lon, lat) %>%
  mutate(type = "city")

all_labels <- bind_rows(station_labels, city_labels)

beach_seine_stations <-
  ggplot() +
  geom_sf(data = sf_regions, fill = "grey95", color = "grey60") +
  geom_sf(data = bay_water_filtered, fill = "lightsteelblue2",
          color = "steelblue3", linewidth = 0.3) +
  geom_sf(data = flowlines_filtered3, color = "steelblue4",
          linewidth = 0.4, alpha = 0.7) +
  geom_sf(data = stations_sf, aes(color = region_label), size = 2) +
  geom_label_repel(
    data = all_labels,
    aes(x = lon, y = lat, label = label,
        size = ifelse(type == "city", 3, 2.5),
        fontface = ifelse(type == "city", "bold.italic", "plain")),
    label.padding = unit(0.15, "lines"),
    max.overlaps = 25, min.segment.length = 0.2,
    box.padding = 0.4, force = 2
  ) +
  scale_size_identity() +
  coord_sf(xlim = c(-122.55, -121.2), ylim = c(37.75, 39.3),
           crs = st_crs(4326), default_crs = st_crs(4326), expand = FALSE) +
  labs(title = "Selected Beach Seine Stations", color = "Region") +
  theme_bw()

ggsave(here("salmon/plots/beach_seine_stations.png"))

```

To retrieve the saved .png file and view it in RStudio...

```{r}
#| label: retrieve-map


library(magick)
image_read(here("salmon/plots/beach_seine_stations.png"))

```
