---
title: "First Flush - Water Quality Analysis Plan"
author: "Dave Bosworth"
date: today
date-format: long
format:
  html:
    toc: true
    toc-expand: true
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| include: false
# Load packages
library(tidyverse)
library(viridisLite)
library(sf)
library(mapview)
library(DT)
library(here)
library(conflicted)

# Declare package conflict preferences
conflicts_prefer(dplyr::filter())

# Load necessary targets
targets::tar_load(c(station_metadata, por_metadata))

# Source necessary functions
source(here("functions/shared/utils_convert_fct_cwq.R"))
```

# WQ Hypotheses

1. Turbidity will increase faster and higher during a first flush event compared to later storm events.
2. Salinity will decrease more dramatically during first flush compared with other storms.
3. Dissolved oxygen will decrease during first flush but recover afterwards.
4. Chlorophyll will decrease more during first flush and stays low during later storm events.
5. Do we need a water temperature hypothesis
6. Other parameters we are interested in?

# Stations

## Sacramento River Mainstem

- USGS Sacramento River at Freeport (11447650)
- DWR-CEMP Sacramento River at Hood (SRH)
- USGS Sacramento River below Georgiana Slough (11447905)
- DWR-NCRO (WQES) Miner Slough near Sacramento River (B9147000) - water quality
- DWR-NCRO (Flow) Miner Slough near Sacramento River (B91470Q) - discharge and velocity
- DWR-NCRO (WQES) Steamboat Slough near Sacramento River (B9145000) - water quality
- DWR-NCRO (Flow) Steamboat Slough near Sacramento River (B91450Q) - discharge and velocity

## Sacramento River DWSC

- USGS Sacramento River Deep Water Ship Channel near Rio Vista (11455335) - only from 2008-2021
- USGS Sacramento River Deep Water Ship Channel near Courtland (11455142)
- USGS RYI and RYF combined
    - USGS Cache Slough at Ryer Island (11455350) - from 2003 to 9/5/2018
    - USGS Cache Slough above Ryer Island Ferry near Rio Vista CA (11455385) - from 9/6/2018 to 2025

## Cache Slough and Liberty Island

- USGS Cache Slough near Hastings Tract near Rio Vista CA (11455280)
- USGS Cache Slough at South Liberty Island near Rio Vista CA (11455315)

## North and South Forks Mokelumne River

- DWR-NCRO (WQES) Mokelumne River Near Hwy 12 (MOK) - water quality
- USGS Mokelumne River at Andrus Island near Terminous CA (11336930) - discharge and velocity
- USGS Little Potato Slough at Terminous CA (11336790)
- USGS North Mokelumne near Walnut Grove CA (11336685)
- USGS South Mokelumne River at New Hope Bridge near Walnut Grove CA (11336680)

## South Delta

- DWR-CEMP San Joaquin River at Prisoners Point (PPT) - water quality
- USGS San Joaquin River at Prisoners Point near Terminous (11313460) - discharge and velocity
- DWR-NCRO (WQES) Old River at Quimby Island (ORQ) - water quality
- USGS Old River at Quimby Island near Bethel Island (11313434) - discharge and velocity
- DWR-NCRO (WQES) Middle River near Holt (HLT) - water quality
- USGS Middle River near Holt (11312685) - discharge and velocity
- DWR-NCRO (WQES) Old River near Bacon Island at USGS Pile (OBI) - water quality
- USGS Old River at Bacon Island (11313405) - discharge and velocity
- USGS Middle River at Middle River (11312676)
- USGS Old River near Byron (11313315)
- DWR-NCRO (WQES) Victoria Canal near Byron (VCU) - water quality
- USGS Victoria Canal near Byron (11312672) - discharge and velocity
- DWR-NCRO (WQES) Old River at Tracy Wildlife Association (TWA) - water quality
- DWR-NCRO (Flow) Old River near Tracy Blvd Bridge (B95380Q) - discharge and velocity
- DWR-CEMP San Joaquin River at Mossdale Bridge (MSD) - water quality
- DWR-NCRO (Flow) San Joaquin River at Mossdale (B95820Q) - discharge and velocity
- USGS San Joaquin River below Garwood Bridge at Stockton (11304810)

## Confluence

- DWR-CEMP Sacramento River at Rio Vista Bridge (RVB) - water quality
- USGS Sacramento River at Rio Vista (11455420) - discharge and velocity
- USGS San Joaquin River at Jersey Point (11337190)
- DWR-CEMP San Joaquin River at Antioch Ship Canal (ANH)
- DWR-CEMP Sacramento River at Chipps Island (MAL)

## Suisun Marsh and Montezuma Slough

- USGS First Mallard Branch near Fairfield CA (381142122015801)
- DWR-WQA Montezuma Slough at National Steel (S-64) - water quality
- DWR-NCRO (Flow) Montezuma Slough at National Steel (E33670Q) - discharge and velocity
- DWR-WQA Montezuma Slough near Beldons Landing (S-49)
- DWR-WQA Suisun Slough at Godfather (S-37)

## Suisun and Honker Bays

- DWR-CEMP Honker Bay near Wheeler Point (HON)
- DWR-CEMP Suisun Bay off Middle Point near Nichols (RYC)
- DWR-CEMP Grizzly Bay at Dolphin near Suisun Slough (GZL)

## San Pablo Bay and Carquinez Strait

- DWR-CEMP Suisun Bay at Martinez (MRZ)
- USGS Suisun Bay at Benicia Bridge near Benicia CA (11455780)
- USGS Carquinez Strait at Carquinez Bridge near Crockett CA (11455820)

## Out of bounds

- DWR-CEMP San Joaquin River McCune Station near Vernalis (SJR) - water quality
- USGS San Joaquin River near Vernalis (11303500) - discharge, water temperature (1972-1998), and SSC
- *DWR-O&M Clifton Court Forebay (CLC)?*
    - Daily Water Temperature: 1988-current
    - Daily Specific Conductance: 1987-current
    - Daily Turbidity (NTU): 1988-current
    - Daily Chlorophyll Fluorescence: 1988-current

# Parameters

For now, we'll look at the following parameters:

- Water Temperature
- Specific Conductance
- Turbidity
- Suspended Sediment Concentration
- Tidally-filtered Discharge
- Discharge (only for SJR)
- Tidally-filtered Velocity

# Map

```{r}
# Convert stations metadata to sf object to be used in map
sf_stations <- station_metadata |>
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326
  )

# Add info for CLC station
# sf_clc <-
#   tibble(
#     Parameter = c(
#       "Water Temperature",
#       "Specific Conductance",
#       "Turbidity (NTU)",
#       "Chlorophyll Fluorescence"
#     ),
#     Start = ymd(c("1988-03-04", "1987-01-01", "1988-03-10", "1988-03-07"))
#   ) |>
#   expand_grid(
#     Source = "DWR-O&M",
#     Station_ID = "CLC",
#     Station_Name = "Clifton Court Forebay",
#     Latitude = 37.829800,
#     Longitude = -121.557400,
#     Interval = "Daily",
#     End = ymd(today())
#   ) |>
#   st_as_sf(
#     coords = c("Longitude", "Latitude"),
#     crs = 4326,
#     sf_column_name = "geom"
#   )

# Import Strata polygons
sf_strata <-
  read_sf(
    here("data/processed/spatial/first_flush_spatial_data.gpkg"),
    layer = "design_strata"
  ) |>
  filter(
    !Stratum %in%
      c("Out of Bounds", "Napa River", "Central SF Bay", "South SF Bay")
  )

sf_stations |>
  mapview(label = "station_abbr", layer.name = "Stations", cex = 5) +
  mapview(
    sf_strata,
    zcol = "Stratum",
    label = "Stratum",
    layer.name = "Strata",
    alpha.regions = 0.4,
    legend = FALSE
  )
```

# Station Metadata

```{r}
# Prepare station metadata to join to period of record metadata
station_metadata_c <- station_metadata |>
  distinct(station_abbr, stratum) |>
  mutate(
    station_abbr = convert_fct_station_abbr(
      replace_values(station_abbr, "RYI" ~ "RYI-RYF")
    )
  ) |>
  drop_na(station_abbr)

# Add strata to period of record metadata and produce interactive DT table
por_metadata |>
  left_join(station_metadata_c, by = join_by(station_abbr)) |>
  relocate(stratum, .after = station_abbr) |>
  mutate(parameter = factor(parameter)) |>
  datatable(filter = "top")
```

# Data Exploration

1. Visual exploration of WQ time-series in relation to storm events at select stations

    a. Turbidity - most stations started around 2010
        i. Increases in relation to storm events.
        ii. Is it typically highest after first event of the year? *Maybe visualize the rate of change or difference in addition to raw values.*
        iii. How does turbidity increase downstream of SR at Freeport? Similar patterns? Is there a lag in response?
        iv. Sacramento River vs. San Joaquin River - how do they compare? How often are storms occurring system-wide vs. in just one watershed?
    b. Salinity
    c. Dissolved Oxygen
    d. Chlorophyll

2. Other visuals besides time-series plots - boxplots?

# Metrics

1. Turbidity
    a. Turbidity maxima after the onset of each storm
    b. Days between first day of storm and turbidity maxima (could help define time lags at downstream stations)
    c. Temporal rate of increase or raw difference (~1 week before storm vs. during storm at its peak)
    d. Spatial extent of turbidity increase (maybe use representative stations in each region to identify whether it increased in that area). Summarize this in some useful way.
    e. Gradient calculations between SR and SJR influenced stations
    f. Compare before (~1 week before storm) to during/after (~1 week after storm) for each storm grouped by region
    g. Some metrics to look at duration of elevated turbidity events in relation to storm events
2. Salinity
    a. Compare before (~1 week before storm) to during/after (~1 week after storm) for each storm grouped by region
3. Dissolved Oxygen
    a. Compare before (~1 week before storm) to during/after (~1 week after storm) for each storm grouped by region
4. Chlorophyll
    a. Compare before (~1 week before storm) to during/after (~1 week after storm) for each storm grouped by region

# Statistical Analyses

1. Turbidity
2. Salinity
3. Dissolved Oxygen
4. Chlorophyll
